**This is only a basic Mini-Project Template.**
**Delete this note and customize formatting according to your preferences and**
**the Mini-Project instructions. Your document must start the `---` below**
---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: ""
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

Remember that code blocks look like this: 

```{r}
#| label: task1-setup
#| message: false
#| warning: false

library(tidyverse)
library(tidycensus)
library(glue)
library(sf)
library(readxl)

# cache tigris shapefile downloads across sessions
options(tigris_use_cache = TRUE)

# project-level data directory (same pattern as mp01, mp02)
data_dir <- file.path("data", "mp03")
if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
}
```


```{r}
#| label: task1-pull-acs1
#| message: false

state_pop_file <- file.path(data_dir, "state_population_2015_2024.rds")

if(!file.exists(state_pop_file)){
    
    # ACS-1 was not released for 2020 (pandemic disruption)
    YEARS <- setdiff(2015:2024, 2020)
    
    state_pop_raw <- map(YEARS, function(yy){
        message(glue("Pulling ACS-1 state population for {yy}"))
        get_acs(
            geography = "state", 
            variables = c(total_population = "B01003_001"),
            year      = yy, 
            survey    = "acs1", 
            geometry  = TRUE,    # grab shapefile too
            cache_table = TRUE   # cache the raw table locally
        ) |> 
            mutate(year = yy)
    }) |> bind_rows()
    
    saveRDS(state_pop_raw, state_pop_file)
} else {
    state_pop_raw <- readRDS(state_pop_file)
}
```


```{r}
#| label: task1-tidy

state_pop <- state_pop_raw |> 
    select(GEOID, 
           state_name       = NAME, 
           total_population = estimate, 
           year, 
           geometry)

# quick sanity check: one row per state per year
state_pop |> 
    st_drop_geometry() |> 
    count(year)
```

```{r}
#| label: task2-function
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(glue)

state_to_state_migration_2024 <- function(){
    
    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }
    
    file_url  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                        "tables/geographic-mobility/2024/state-to-state-migration/",
                        "State_to_State_Migration_Table_2024_T13.xlsx")
    file_path <- file.path(data_dir, "State_to_State_Migration_Table_2024.xlsx")
    
    if(!file.exists(file_path)){
        message(glue("Downloading 2024 state-to-state migration file from {file_url}"))
        download.file(file_url, destfile = file_path, quiet = TRUE, mode = "wb")
    }
    
    # off-diagonal flows from the "Table" sheet
    flows_table <- read_excel(
            file_path,
            sheet     = "Table",
            skip      = 8,
            col_names = c("state_current", "state_1y", "estimate", "moe")
        ) |>
        mutate(state_current = str_trim(state_current),
               state_1y      = str_trim(state_1y)) |>
        filter(state_current %in% state.name,
               state_1y      %in% state.name) |>
        filter(estimate != "X") |>
        mutate(estimate = if_else(estimate == "N", "0", estimate),
               estimate = as.integer(estimate)) |>
        # convert full names to two-letter abbreviations
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)]) |>
        select(state_current, state_1y, population = estimate) |>
        mutate(year_current = 2024L, year_1y = 2023L)
    
    # stayed-in-state diagonal from the Supplemental sheet:
    # col 4 = same residence a year ago, col 6 = moved within same state
    flows_stayed <- read_excel(
            file_path,
            sheet     = "Supplemental - Current Res",
            skip      = 8,
            col_names = FALSE
        ) |>
        select(state_current  = 1,
               same_residence = 4,
               mover_in_state = 6) |>
        mutate(state_current = str_trim(state_current)) |>
        filter(state_current %in% state.name) |>
        mutate(population = as.integer(same_residence) +
                            as.integer(mover_in_state),
               state_1y   = state_current) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)],
               year_current  = 2024L,
               year_1y       = 2023L) |>
        select(state_current, state_1y, population, year_current, year_1y)
    
    bind_rows(flows_table, flows_stayed) |>
        arrange(state_current, state_1y)
}
```

```{r}
#| label: task2-check

flows_2024 <- state_to_state_migration_2024()

# Shape checks
dim(flows_2024)                         # expect 2500 x 5
n_distinct(flows_2024$state_current)    # expect 50
n_distinct(flows_2024$state_1y)         # expect 50

# Spot-check vs raw Excel:
# Table sheet row 19: AL <- FL = 16,423
flows_2024 |> filter(state_current == "AL", state_1y == "FL")

# Supplemental row for Alabama: 4,522,731 + 437,829 = 4,960,560 stayed-in-AL
flows_2024 |> filter(state_current == "AL", state_1y == "AL")

flows_2024 |> head(5)
```


```{r}
#| label: task3-step2-download

file_url_2023  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                         "tables/geographic-mobility/2023/state-to-state-migration/",
                         "State_to_State_Migration_Table_2023_T13.xlsx")
file_path_2023 <- file.path(data_dir, "State_to_State_Migration_Table_2023.xlsx")

if(!file.exists(file_path_2023)){
    message("Downloading 2023 state-to-state migration file")
    download.file(file_url_2023,
                  destfile = file_path_2023,
                  quiet    = TRUE,
                  mode     = "wb")
}
```

```{r}
#| label: task3-step3-read

# The row/column state ordering in the 2023 sheet matches alphabetical
# with DC inserted after Delaware.
states_with_dc <- c(state.name[1:8], "District of Columbia", state.name[9:50])

# (a) First 6 columns: col D = same residence (didn't move), col F = moved
# within same state. Their sum gives total who stayed inside the state.
# Rows 12-62 cover the 50 states + DC.
stayed_2023 <- read_excel(
        file_path_2023,
        range     = "A12:F62",
        col_names = c("state_current", "pop_est", "pop_moe",
                      "same_res",      "same_res_moe", "same_state")
    ) |>
    mutate(state_current = str_trim(state_current)) |>
    filter(state_current %in% state.name) |>                  # drop DC -> 50 rows
    mutate(population   = as.integer(same_res) + as.integer(same_state),
           state_1y     = state_current,
           year_current = 2023L,
           year_1y      = 2022L) |>
    select(state_current, state_1y, population, year_current, year_1y)

# (b) Bulk flow matrix J:DG. Columns alternate Estimate/MOE, so interleaving
# the state names with NA gives proper names to the Estimate columns and
# placeholder names to the MOE columns, which any_of() then drops.
between_2023 <- read_excel(
        file_path_2023,
        range     = "J12:DG62",
        col_names = vec_interleave(states_with_dc, NA)
    ) |>
    select(any_of(states_with_dc)) |>           # keep 51 Estimate cols
    mutate(state_current = states_with_dc) |>   # tag rows with destination
    pivot_longer(cols      = -state_current,
                 names_to  = "state_1y",
                 values_to = "population") |>
    filter(state_current != "District of Columbia",
           state_1y      != "District of Columbia") |>   # drop DC rows/cols
    filter(state_current != state_1y) |>                 # drop diagonals ("N/A")
    mutate(population   = as.integer(population),
           year_current = 2023L,
           year_1y      = 2022L) |>
    select(state_current, state_1y, population, year_current, year_1y)
```

```{r}
#| label: task3-step4-combine

flows_2023 <- bind_rows(stayed_2023, between_2023) |>
    arrange(state_current, state_1y)

dim(flows_2023)                         # 2500 x 5
n_distinct(flows_2023$state_current)    # 50
n_distinct(flows_2023$state_1y)         # 50

# Spot checks against the raw spreadsheet:
# Row 13 (Alaska) Col J (Alabama) = 204  ->  people in AK who lived in AL a year ago
flows_2023 |> filter(state_current == "Alaska", state_1y == "Alabama")

# Alabama diagonal: 4,482,992 (same res) + 437,745 (moved within AL) = 4,920,737
flows_2023 |> filter(state_current == "Alabama", state_1y == "Alabama")

flows_2023 |> head()
```

```{r}
#| label: task3-function
#| message: false
#| warning: false

library(vctrs)   # for vec_interleave()

state_to_state_migration_2023 <- function(){

    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }

    file_url  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                        "tables/geographic-mobility/2023/state-to-state-migration/",
                        "State_to_State_Migration_Table_2023_T13.xlsx")
    file_path <- file.path(data_dir, "State_to_State_Migration_Table_2023.xlsx")
    if(!file.exists(file_path)){
        message("Downloading 2023 state-to-state migration file")
        download.file(file_url, destfile = file_path, quiet = TRUE, mode = "wb")
    }

    states_with_dc <- c(state.name[1:8], "District of Columbia", state.name[9:50])

    # stayed-in-state diagonal (col D same residence + col F moved within state)
    stayed <- read_excel(
            file_path,
            range     = "A12:F62",
            col_names = c("state_current", "pop_est", "pop_moe",
                          "same_res",      "same_res_moe", "same_state")
        ) |>
        mutate(state_current = str_trim(state_current)) |>
        filter(state_current %in% state.name) |>
        mutate(population   = as.integer(same_res) + as.integer(same_state),
               state_1y     = state_current,
               year_current = 2023L,
               year_1y      = 2022L) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)]) |>
        select(state_current, state_1y, population, year_current, year_1y)

    # between-state flow matrix J:DG; interleave state names with NA so the
    # Estimate columns are named and MOE columns get dropped by any_of()
    between <- read_excel(
            file_path,
            range     = "J12:DG62",
            col_names = vec_interleave(states_with_dc, NA)
        ) |>
        select(any_of(states_with_dc)) |>
        mutate(state_current = states_with_dc) |>
        pivot_longer(cols      = -state_current,
                     names_to  = "state_1y",
                     values_to = "population") |>
        filter(state_current != "District of Columbia",
               state_1y      != "District of Columbia") |>
        filter(state_current != state_1y) |>
        mutate(population   = as.integer(population),
               year_current = 2023L,
               year_1y      = 2022L) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)]) |>
        select(state_current, state_1y, population, year_current, year_1y)

    bind_rows(stayed, between) |>
        arrange(state_current, state_1y)
}
```

```{r}
#| label: task4-function

state_to_state_migration <- function(){
    bind_rows(
        state_to_state_migration_2024(),
        state_to_state_migration_2023()
    ) |>
        arrange(year_current, state_current, state_1y)
}
```

```{r}
#| label: task4-check

flows <- state_to_state_migration()

dim(flows)                                    # expect 5000 x 5
n_distinct(flows$state_current)               # 50
n_distinct(flows$state_1y)                    # 50

# match the four example rows exactly
flows |> filter(state_current == "AL", state_1y == "AL", year_current == 2024)  # 4,960,560
flows |> filter(state_current == "AL", state_1y == "AK", year_current == 2024)  # 387
flows |> filter(state_current == "AL", state_1y == "AL", year_current == 2023)  # 4,920,737
flows |> filter(state_current == "AL", state_1y == "AK", year_current == 2023)  # 780
```

```{r}
#| label: task5-step1-dir
#| message: false
#| warning: false

library(tidyverse)
library(jsonlite)

data_dir <- file.path("data", "mp03")
if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
}
```

```{r}
#| label: task5-step2-download

# 2016-2020 ACS-5 metro-to-metro flows (MOVEDIN = movers into FULL1 from FULL2,
# MOVEDOUT = movers into FULL2 from FULL1). MSA summary level code is 310;
# the API predicate for MSA/micropolitan is spelled out below and URL-encoded
# via URLencode() so the %20 and / make it through cleanly.
api_path  <- paste0(
    "https://api.census.gov/data/2020/acs/flows",
    "?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME",
    "&for=", URLencode("metropolitan statistical area/micropolitan statistical area:*")
)
json_path <- file.path(data_dir, "metro_to_metro_flows_2016_2020.json")

if(!file.exists(json_path)){
    message("Downloading 2016-2020 metro-to-metro migration flows from Census API")
    download.file(api_path, destfile = json_path, quiet = TRUE, mode = "wb")
}
```

```{r}
#| label: task5-step3-read

metro_raw <- read_json(json_path, simplifyVector = TRUE)
# returns a matrix: first row = column names, rest = data rows
dim(metro_raw)
metro_raw[1:3, ]
```

```{r}
#| label: task5-step4-fix-headers

move_row_to_colnames <- function(X, row = 1){
    X    <- as_tibble(X)
    Xrow <- X[row, ]
    X    <- X[-row, ]
    colnames(X) <- Xrow
    readr::type_convert(X)
}

metro_typed <- move_row_to_colnames(metro_raw)
glimpse(metro_typed)
```

```{r}
#| label: task5-step5-clean-columns

metro_clean <- metro_typed |>
    select(metro_current = FULL1_NAME,
           metro_1y      = FULL2_NAME,
           moved_in      = MOVEDIN,
           moved_out     = MOVEDOUT)
```

```{r}
#| label: task5-step6-pull-state

pull_state_from_metro <- function(metro_name){
    str_extract(metro_name,
                ".*, (\\S{2})[-[:alpha:]]* Metro Area",
                group = 1)
}

metro_clean <- metro_clean |>
    mutate(state_current = pull_state_from_metro(metro_current),
           state_1y      = pull_state_from_metro(metro_1y))
```

```{r}
#| label: task5-function

metro_to_metro_migration <- function(){

    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }

    api_path  <- paste0(
        "https://api.census.gov/data/2020/acs/flows",
        "?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME",
        "&for=", URLencode(
            "metropolitan statistical area/micropolitan statistical area:*")
    )
    json_path <- file.path(data_dir, "metro_to_metro_flows_2016_2020.json")

    if(!file.exists(json_path)){
        message("Downloading 2016-2020 metro-to-metro migration flows from Census API")
        download.file(api_path, destfile = json_path, quiet = TRUE, mode = "wb")
    }

    move_row_to_colnames <- function(X, row = 1){
        X    <- as_tibble(X)
        Xrow <- X[row, ]
        X    <- X[-row, ]
        colnames(X) <- Xrow
        readr::type_convert(X)
    }

    pull_state_from_metro <- function(metro_name){
        str_extract(metro_name,
                    ".*, (\\S{2})[-[:alpha:]]* Metro Area",
                    group = 1)
    }

    read_json(json_path, simplifyVector = TRUE) |>
        move_row_to_colnames() |>
        select(metro_current = FULL1_NAME,
               metro_1y      = FULL2_NAME,
               moved_in      = MOVEDIN,
               moved_out     = MOVEDOUT) |>
        mutate(state_current = pull_state_from_metro(metro_current),
               state_1y      = pull_state_from_metro(metro_1y))
}
```

```{r}
#| label: task5-check

metro_flows <- metro_to_metro_migration()

dim(metro_flows)
glimpse(metro_flows)

# primary-state extraction worked?
metro_flows |>
    distinct(metro_current, state_current) |>
    slice_sample(n = 8)

# how many rows don't match the "Metro Area" regex
#   (these are "Outside Metro Area ..." catch-all origins,
#   or origins outside the US — `state_current`/`state_1y` will be NA for them)
metro_flows |> summarise(na_current = sum(is.na(state_current)),
                         na_1y      = sum(is.na(state_1y)))
```
or inline, like this: $x$ is `r x`.

## Exploratory Analysis

```{r}
#| label: task6-setup
#| message: false
#| warning: false

library(tidyverse)
library(gt)
library(sf)

# NY state abbreviation shorthand
NY <- "NY"

# NYC metro — the full MSA string as it appears in metro_flows
NYC_METRO <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"
```

```{r}
#| label: q1-growth-rates

state_growth <- state_pop |>
    st_drop_geometry() |>
    filter(year %in% c(2015, 2024)) |>
    select(state_name, year, total_population) |>
    pivot_wider(names_from = year, values_from = total_population,
                names_prefix = "pop_") |>
    mutate(growth_rate = (pop_2024 - pop_2015) / pop_2015) |>
    arrange(desc(growth_rate))

state_growth |>
    slice_head(n = 10) |>
    gt() |>
    tab_header(title = "Top 10 States by Population Growth, 2015 → 2024") |>
    fmt_number(columns = c(pop_2015, pop_2024), decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = growth_rate, decimals = 1) |>
    cols_label(state_name = "State",
               pop_2015   = "2015 Population",
               pop_2024   = "2024 Population",
               growth_rate = "Growth Rate")
```

Idaho, Utah, Texas, Florida, and the rest of the Mountain West / South dominate the top — the well-documented 2010s Sun Belt migration pattern. I'd expect West Virginia, Illinois, Louisiana, and Mississippi near the bottom (shrinking populations).

```{r}
#| label: q2-ny-state-flows

flows_latest <- flows |> filter(year_current == 2024)

# Inflows: people in NY in 2024 who were elsewhere in 2023
into_ny <- flows_latest |>
    filter(state_current == NY, state_1y != NY) |>
    arrange(desc(population)) |>
    slice_head(n = 10) |>
    select(state_1y, population)

# Outflows: people elsewhere in 2024 who were in NY in 2023
out_of_ny <- flows_latest |>
    filter(state_1y == NY, state_current != NY) |>
    arrange(desc(population)) |>
    slice_head(n = 10) |>
    select(state_current, population)

into_ny |> gt() |>
    tab_header(title = "Top 10 Origins of People Who Moved TO New York (2023 → 2024)") |>
    fmt_number(columns = population, decimals = 0, use_seps = TRUE) |>
    cols_label(state_1y = "Previous State", population = "People")

out_of_ny |> gt() |>
    tab_header(title = "Top 10 Destinations of People Who Moved FROM New York (2023 → 2024)") |>
    fmt_number(columns = population, decimals = 0, use_seps = TRUE) |>
    cols_label(state_current = "New State", population = "People")
```

Expected pattern: the biggest flow in both directions is New Jersey (commuter/metro overflow), followed by Florida, California, Pennsylvania, Texas, Connecticut, Massachusetts. The out-flows to FL and TX should dwarf the in-flows from those places.

```{r}
#| label: q3-nyc-metro-flows

into_nyc <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y      != NYC_METRO,
           !is.na(moved_in)) |>
    arrange(desc(moved_in)) |>
    slice_head(n = 10) |>
    select(metro_1y, moved_in)

out_of_nyc <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y      != NYC_METRO,
           !is.na(moved_out)) |>
    arrange(desc(moved_out)) |>
    slice_head(n = 10) |>
    select(metro_1y, moved_out)

into_nyc |> gt() |>
    tab_header(title = "Top 10 Origin Metros — Moving INTO NYC Metro",
               subtitle = "2016–2020 ACS-5 estimates") |>
    fmt_number(columns = moved_in, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Origin", moved_in = "People")

out_of_nyc |> gt() |>
    tab_header(title = "Top 10 Destination Metros — Moving OUT OF NYC Metro",
               subtitle = "2016–2020 ACS-5 estimates") |>
    fmt_number(columns = moved_out, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Destination", moved_out = "People")
```

A quick note on dates: the metro-to-metro data is the 2016–2020 ACS-5 window — the most recent MSA-level flow file the Census Bureau publishes — so this describes pre- and early-pandemic patterns, not 2023–2024. The top origins and destinations are likely Miami-Fort Lauderdale, Los Angeles, Philadelphia, Boston, Washington DC, Tampa, Orlando, Poughkeepsie.

```{r}
#| label: q4-state-migration-totals

state_net <- flows_latest |>
    filter(state_current != state_1y) |>     # drop diagonals (non-movers)
    group_by(state_current) |>
    summarise(in_migration  = sum(population), .groups = "drop") |>
    rename(state = state_current) |>
    full_join(
        flows_latest |>
            filter(state_current != state_1y) |>
            group_by(state_1y) |>
            summarise(out_migration = sum(population), .groups = "drop") |>
            rename(state = state_1y),
        by = "state"
    ) |>
    mutate(net_migration = in_migration - out_migration) |>
    arrange(desc(net_migration))

# Top and bottom 10 net migration
bind_rows(
    state_net |> slice_head(n = 10) |> mutate(group = "Top 10"),
    state_net |> slice_tail(n = 10) |> mutate(group = "Bottom 10")
) |>
    gt(groupname_col = "group") |>
    tab_header(title = "State-Level Migration, 2023 → 2024",
               subtitle = "Sum of interstate flows only (intl. migration excluded)") |>
    fmt_number(columns = c(in_migration, out_migration, net_migration),
               decimals = 0, use_seps = TRUE) |>
    cols_label(state = "State",
               in_migration  = "In",
               out_migration = "Out",
               net_migration = "Net")
```

The net migration ranking captures which states are gaining vs. losing residents through interstate movement. Top gainers will be TX, FL, NC, SC, TN, AZ; biggest losers will be CA, NY, IL, NJ, MA — the coastal-to-Sun Belt trend.

```{r}
#| label: q5-metro-migration-totals

# For each metro (FULL1 = current), MOVEDIN is people arriving from FULL2,
# MOVEDOUT is people leaving to FULL2. Sum over all FULL2 partners.
metro_totals <- metro_flows |>
    filter(metro_current != metro_1y) |>
    group_by(metro_current) |>
    summarise(in_migration  = sum(moved_in,  na.rm = TRUE),
              out_migration = sum(moved_out, na.rm = TRUE),
              .groups = "drop") |>
    mutate(net_migration = in_migration - out_migration) |>
    # restrict to actual US metros (regex matched a 2-letter state code)
    filter(!is.na(str_extract(metro_current, ", \\S{2}[-[:alpha:]]* Metro Area"))) |>
    arrange(desc(net_migration))

bind_rows(
    metro_totals |> slice_head(n = 10) |> mutate(group = "Top 10 (largest net gain)"),
    metro_totals |> slice_tail(n = 10) |> mutate(group = "Bottom 10 (largest net loss)")
) |>
    gt(groupname_col = "group") |>
    tab_header(title = "Metro-Level Migration Totals",
               subtitle = "2016–2020 ACS-5 estimates, sum over all partner metros") |>
    fmt_number(columns = c(in_migration, out_migration, net_migration),
               decimals = 0, use_seps = TRUE) |>
    cols_label(metro_current = "Metro Area",
               in_migration  = "In",
               out_migration = "Out",
               net_migration = "Net")
```

Metro-level winners track the state story — expect the Dallas, Houston, Phoenix, Atlanta, Tampa, and Austin metros near the top; the LA, NYC, Chicago, DC, Boston metros near the bottom for net migration.

```{r}
#| label: q6-stayer-fraction

stayer_frac <- flows_latest |>
    filter(state_current == state_1y) |>           # diagonal rows only
    select(state = state_current, stayers = population) |>
    left_join(
        state_pop |>
            st_drop_geometry() |>
            filter(year == 2024) |>
            mutate(state = state.abb[match(state_name, state.name)]) |>
            select(state, total_pop = total_population),
        by = "state"
    ) |>
    filter(!is.na(total_pop)) |>
    mutate(stayer_rate = stayers / total_pop) |>
    arrange(desc(stayer_rate))

stayer_frac |>
    slice_head(n = 10) |>
    gt() |>
    tab_header(title = "Top 10 States by Stayer Rate",
               subtitle = "Share of residents who lived in the same state one year earlier") |>
    fmt_number(columns = c(stayers, total_pop), decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = stayer_rate, decimals = 2) |>
    cols_label(state = "State",
               stayers     = "Same-State Stayers (2024)",
               total_pop   = "Total Population (2024)",
               stayer_rate = "Stayer Rate")
```

The stayer count comes from the "same state one year ago" diagonal (from the Supplemental sheet, summing same_residence + moved_within_state), and the denominator is the ACS-1 2024 total population. Small universe mismatch — stayers are the population 1 year and over, the denominator includes infants under 1 — but the effect on the ranking is negligible. Expect Michigan, Texas, Louisiana, Ohio (low out-migration states with weak net mobility) near the top.

```{r}
#| label: q7-migration-share-of-growth

pop_change <- state_pop |>
    st_drop_geometry() |>
    filter(year %in% c(2023, 2024)) |>
    select(state_name, year, total_population) |>
    pivot_wider(names_from = year, values_from = total_population,
                names_prefix = "pop_") |>
    mutate(state = state.abb[match(state_name, state.name)],
           pop_change = pop_2024 - pop_2023) |>
    select(state, pop_change) |>
    filter(!is.na(state))

migration_share <- pop_change |>
    left_join(state_net |> select(state, net_migration), by = "state") |>
    mutate(migration_share = net_migration / pop_change) |>
    # only interpretable where population actually grew
    filter(pop_change > 0) |>
    arrange(desc(migration_share))

migration_share |>
    slice_head(n = 10) |>
    gt() |>
    tab_header(title = "States Where Growth Is Most Migration-Driven",
               subtitle = "Net interstate migration as a share of total 2023 → 2024 population growth") |>
    fmt_number(columns = c(pop_change, net_migration), decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = migration_share, decimals = 0) |>
    cols_label(state           = "State",
               pop_change      = "Total Population Growth",
               net_migration   = "Net Interstate Migration",
               migration_share = "Share from Migration")
```

States where the "Share from Migration" number is under 100% are reading the statistic the intuitive way: migration is the dominant component of growth, with births-minus-deaths and international migration rounding out the picture. New Hampshire (96%), Maine (91%), Oklahoma (82%), North Dakota (60%), Nevada (55%), South Carolina (52%), and Alabama (46%) all fit this pattern — they grew by a sensible margin, and interstate migration explains roughly half to nearly all of it.

Shares above 100% look surprising but have a clear demographic interpretation: interstate migration is adding more people than the state's total growth, which means the other components of population change (natural increase + international migration) are net-negative. Vermont is the most extreme case: its total population grew by only 1,029 people year over year, but +9,231 Americans moved in on net — the missing ~8,200 people is roughly what the state lost to deaths exceeding births plus whatever small international flow contributed. Vermont has the oldest median age in the country, so negative natural increase isn't a glitch in the data; it's a demographic fact. The same story drives Mississippi (322%) and Wyoming (182%): small, aging populations where interstate migration is a lifeline keeping total growth positive at all.

Ranking by share alone therefore over-weights tiny or aging states. The more substantively meaningful takeaway is that for every state in the top ten, net interstate migration is the dominant source of growth — either because it's the largest positive component (rows 4–10) or because it's compensating for demographic decline (rows 1–3). Either way, this is strong evidence that the continued redistribution of Americans toward the South and Mountain West — and away from the Northeast and upper Midwest — is the primary engine of U.S. state-level population change in 2024, much more so than differential birth rates or international migration.
```{r}
#| label: q8-ny-top-partner

top_into_ny  <- into_ny    |> slice_max(population, n = 1)
top_out_of_ny <- out_of_ny |> slice_max(population, n = 1)

top_into_ny
top_out_of_ny
```

The single state sending the most people to NY in 2023–2024 was `r top_into_ny$state_1y` (about `scales::comma(top_into_ny$population)` people). The single state absorbing the most NY leavers was `top_out_of_ny$state_current` (about `scales::comma(top_out_of_ny$population)` people).

```{r}
#| label: q9-nyc-top-nonny-partner

# helper: extract the first 2-letter state token from a metro name
metro_state <- function(m) str_extract(m, ".*, (\\S{2})[-[:alpha:]]* Metro Area", group = 1)

# NYC's biggest non-NY-primary inbound partner
top_into_nyc_nonny <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y      != NYC_METRO,
           !is.na(moved_in)) |>
    mutate(origin_state = metro_state(metro_1y)) |>
    filter(origin_state != "NY" | is.na(origin_state)) |>
    slice_max(moved_in, n = 5) |>
    select(metro_1y, moved_in)

# NYC's biggest non-NY-primary outbound partner
top_out_of_nyc_nonny <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y      != NYC_METRO,
           !is.na(moved_out)) |>
    mutate(dest_state = metro_state(metro_1y)) |>
    filter(dest_state != "NY" | is.na(dest_state)) |>
    slice_max(moved_out, n = 5) |>
    select(metro_1y, moved_out)

top_into_nyc_nonny |> gt() |>
    tab_header(title = "Top 5 Non-NY Metros Sending People to NYC",
               subtitle = "2016–2020 ACS-5") |>
    fmt_number(columns = moved_in, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Metro", moved_in = "People")

top_out_of_nyc_nonny |> gt() |>
    tab_header(title = "Top 5 Non-NY Metros Receiving NYC Leavers",
               subtitle = "2016–2020 ACS-5") |>
    fmt_number(columns = moved_out, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Metro", moved_out = "People")
```

Background knowledge part: the largest metro in New York State is the NYC metro itself (population ~19M, far larger than Buffalo, Rochester, or Albany). Note that the NYC MSA straddles NY, NJ, and PA — I defined "non-NY" here as metros whose primary (first-listed) state is not NY, so NJ-primary metros count as external even though they're functionally part of the NYC region. If you want me to redefine "outside NY" more strictly (e.g., the whole NY-NJ-PA tri-state), say the word.

```{r}
#| label: q10-nyc-connection
#| fig-width: 11
#| fig-height: 6
#| message: false

library(patchwork)   # for arranging the two charts side by side

# denominators: for each metro X, total inflow from and outflow to all partners
metro_denoms <- metro_flows |>
    filter(metro_current != metro_1y) |>
    group_by(metro_current) |>
    summarise(total_in  = sum(moved_in,  na.rm = TRUE),
              total_out = sum(moved_out, na.rm = TRUE),
              .groups = "drop")

# numerators: for each metro X, flow FROM NYC into X and flow INTO NYC from X.
# The metro flows table reports each ordered pair once, so depending on whether
# NYC shows up as `metro_current` or `metro_1y`, MOVEDIN/MOVEDOUT mean different things:
#   - metro_current == NYC: MOVEDIN = NYC <- other; MOVEDOUT = NYC -> other
#   - metro_1y      == NYC: MOVEDIN = X  <- NYC;   MOVEDOUT = X  -> NYC
nyc_edges <- metro_flows |>
    filter(metro_current != metro_1y,
           (metro_current == NYC_METRO) | (metro_1y == NYC_METRO)) |>
    mutate(other_metro = if_else(metro_current == NYC_METRO, metro_1y, metro_current),
           nyc_to_other = case_when(
               metro_current == NYC_METRO ~ moved_out,
               metro_1y      == NYC_METRO ~ moved_in
           ),
           other_to_nyc = case_when(
               metro_current == NYC_METRO ~ moved_in,
               metro_1y      == NYC_METRO ~ moved_out
           )) |>
    group_by(other_metro) |>
    summarise(nyc_to_other = sum(nyc_to_other, na.rm = TRUE),
              other_to_nyc = sum(other_to_nyc, na.rm = TRUE),
              .groups = "drop")

connection <- nyc_edges |>
    left_join(metro_denoms, by = c("other_metro" = "metro_current")) |>
    mutate(share_of_inflow = nyc_to_other / total_in) |>
    filter(!is.na(str_extract(other_metro, ", \\S{2}[-[:alpha:]]* Metro Area")),
           total_in  >= 1000,
           total_out >= 1000)

# --- Panel 1: overall top 15 (will be dominated by the commuter belt) ---
p1 <- connection |>
    slice_max(share_of_inflow, n = 15) |>
    mutate(metro_short = str_remove(other_metro, " Metro Area")) |>
    ggplot(aes(x = share_of_inflow,
               y = fct_reorder(metro_short, share_of_inflow))) +
    geom_col(fill = "steelblue") +
    geom_text(aes(label = scales::percent(share_of_inflow, accuracy = 0.1)),
              hjust = -0.1, size = 3) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.18))) +
    labs(title    = "Structural ties: NYC commuter belt",
         subtitle = "All metros, including NYC's neighbors",
         x = "Share of inflow from NYC", y = NULL) +
    theme_minimal(base_size = 10)

# --- Panel 2: distant metros only (NY/NJ/CT/PA removed) ---
neighbor_states <- c("NY", "NJ", "CT", "PA")

p2 <- connection |>
    mutate(other_state = str_extract(other_metro, ", (\\S{2})", group = 1)) |>
    filter(!other_state %in% neighbor_states,
           total_in >= 5000) |>                 # bigger floor away from NYC
    slice_max(share_of_inflow, n = 15) |>
    mutate(metro_short = str_remove(other_metro, " Metro Area")) |>
    ggplot(aes(x = share_of_inflow,
               y = fct_reorder(metro_short, share_of_inflow))) +
    geom_col(fill = "#b35806") +
    geom_text(aes(label = scales::percent(share_of_inflow, accuracy = 0.1)),
              hjust = -0.1, size = 3) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.18))) +
    labs(title    = "Long-distance affinities",
         subtitle = "Excludes metros whose primary state is NY, NJ, CT, or PA",
         x = "Share of inflow from NYC", y = NULL) +
    theme_minimal(base_size = 10)

p1 + p2 + plot_annotation(
    title    = "Metros Most Disproportionately Fed by NYC",
    subtitle = "Share of each metro's total incoming migration coming from the NYC MSA (2016–2020 ACS-5)"
)
```

The left panel ranks all metros by NYC's share of their total in-migration, and is unsurprisingly dominated by the NYC commuter belt: Trenton-Princeton (97.6%), Poughkeepsie (92.3%), Bridgeport-Stamford, East Stroudsburg, Kingston, Allentown, Scranton — metros the Census treats as standalone MSAs but which function as NYC's outer suburbs. Any ranking that doesn't filter these out will just describe the geography of the tri-state area, not genuine migration preferences.

The right panel strips out NY/NJ/CT/PA to surface the long-distance affinities — metros far enough from NYC that a high NYC share reflects a real pull rather than physical proximity. This is the cut that matches the professor's Miami example. Florida destinations usually dominate this ranking (Miami-Fort Lauderdale, Cape Coral, Port St. Lucie, North Port-Sarasota, Palm Bay), reflecting the well-documented New York → Florida migration channel. Other metros likely to appear: Los Angeles, Boston, the DC area, and occasionally Atlanta or a North Carolina metro. The right panel is the more substantively interesting read — it tells you which places New Yorkers leave for, separating the real mobility story from the commuting-pattern artifact on the left.

```{r}
#| label: task7-step1-abroad

# US-total movers from abroad — pulled directly from the two cached xlsx files.
# 2024 file: "Supplemental - Current Res" sheet has a consolidated column with this
#   figure; US-total row puts it at (row 9, col J).
# 2023 file: the main "Table" sheet reports it on the US-total row (row 10) in the
#   'Different state of residence - Total' column DH. Footnotes confirm this
#   total covers movers from Puerto Rico, U.S. Island Areas, and foreign countries,
#   matching the 2024 definition of "from abroad".

M_2024 <- read_excel(
        file.path("data", "mp03", "State_to_State_Migration_Table_2024.xlsx"),
        sheet = "Supplemental - Current Res",
        range = "J9",
        col_names = "from_abroad"
    ) |> pull(from_abroad)

M_2023 <- read_excel(
        file.path("data", "mp03", "State_to_State_Migration_Table_2023.xlsx"),
        sheet = "Table",
        range = "DH10",
        col_names = "from_abroad"
    ) |> pull(from_abroad)

M_2024   # movers from abroad during 2023 -> 2024
M_2023   # movers from abroad during 2022 -> 2023
```

```{r}
#| label: task7-step2-growth-rate

# National population totals in each year, summing the 50 state ACS-1 estimates.
# Keeping the frame as just (year, total_pop) so this plays nicely with joins.
national_pop <- state_pop |>
    st_drop_geometry() |>
    filter(year %in% c(2022, 2023, 2024)) |>
    group_by(year) |>
    summarise(national_pop = sum(total_population), .groups = "drop")

P_2022 <- national_pop |> filter(year == 2022) |> pull(national_pop)
P_2023 <- national_pop |> filter(year == 2023) |> pull(national_pop)
P_2024 <- national_pop |> filter(year == 2024) |> pull(national_pop)

# g-bar for the two one-year windows we care about.
# For the 2023 -> 2024 window, M_2024 is the abroad component; denominator is P_2023.
# For the 2022 -> 2023 window, use M_2023 and denominator P_2022.
g_2024 <- (P_2024 - P_2023 - M_2024) / P_2023
g_2023 <- (P_2023 - P_2022 - M_2023) / P_2022

tibble(window = c("2023 -> 2024", "2022 -> 2023"),
       g_bar  = c(g_2024,        g_2023))
```

```{r}
#| label: task7-step3-m-2024

# State populations by (state abbrev, year) for easy joins
state_pop_tidy <- state_pop |>
    st_drop_geometry() |>
    mutate(state = state.abb[match(state_name, state.name)]) |>
    filter(!is.na(state)) |>
    select(state, year, pop = total_population)

# Pair table: P_i^{2023} and P_j^{2023} attached to each (destination i, origin j) flow.
# flows rows where year_current == 2024 represent moves from 2023 -> 2024, so t = 2023.
m_2024 <- flows |>
    filter(year_current == 2024, state_current != state_1y) |>
    left_join(state_pop_tidy |> filter(year == 2023) |> select(state, P_i = pop),
              by = c("state_current" = "state")) |>
    left_join(state_pop_tidy |> filter(year == 2023) |> select(state, P_j = pop),
              by = c("state_1y"      = "state")) |>
    mutate(m = population / sqrt(P_i * P_j)) |>
    select(state_current, state_1y, M_ji = population, P_i, P_j, m)

nrow(m_2024)         # expect 2450
summary(m_2024$m)
head(m_2024)
```

```{r}
#| label: task7-step4-m-2023

m_2023 <- flows |>
    filter(year_current == 2023, state_current != state_1y) |>
    left_join(state_pop_tidy |> filter(year == 2022) |> select(state, P_i = pop),
              by = c("state_current" = "state")) |>
    left_join(state_pop_tidy |> filter(year == 2022) |> select(state, P_j = pop),
              by = c("state_1y"      = "state")) |>
    mutate(m = population / sqrt(P_i * P_j)) |>
    select(state_current, state_1y, M_ji = population, P_i, P_j, m)

nrow(m_2023)
summary(m_2023$m)
```

```{r}
#| label: task7-step5-average-params

# Average g-bar
g_bar <- mean(c(g_2024, g_2023))
g_bar

# Average m for each state pair. Keep (state_current, state_1y) as the key
# and take the simple mean of the two m's (equal weight, per the task).
m_avg <- m_2024 |> select(state_current, state_1y, m_2024 = m) |>
    left_join(
        m_2023 |> select(state_current, state_1y, m_2023 = m),
        by = c("state_current", "state_1y")
    ) |>
    mutate(m = (m_2024 + m_2023) / 2) |>
    select(state_current, state_1y, m)

nrow(m_avg)            # 2450
head(m_avg)
```

```{r}
#| label: task7-step6-projection

# One-step update: given a tibble of (state, pop) for year t, return (state, pop) for t+1.
# P_i^{t+1} = P_i^{t} * (1 + g_bar) + sum_{j != i} m_{j -> i} * sqrt(P_i^{t} * P_j^{t})
step_forward <- function(pop_t, m_matrix, g){
    # migration contribution: for each destination i, sum over origins j.
    # In our m_avg, state_current = i (destination), state_1y = j (origin).
    mig_contrib <- m_matrix |>
        left_join(pop_t |> rename(P_i = pop), by = c("state_current" = "state")) |>
        left_join(pop_t |> rename(P_j = pop), by = c("state_1y"      = "state")) |>
        mutate(term = m * sqrt(P_i * P_j)) |>
        group_by(state = state_current) |>
        summarise(mig_in = sum(term), .groups = "drop")
    
    pop_t |>
        left_join(mig_contrib, by = "state") |>
        mutate(pop = pop * (1 + g) + mig_in) |>
        select(state, pop)
}

# Starting point: 2024 populations (50 states)
pop_2024 <- state_pop_tidy |>
    filter(year == 2024) |>
    select(state, pop)

# Iterate 2024 -> 2030 (6 steps), storing each year in a long tibble
projection <- pop_2024 |> mutate(year = 2024)

current <- pop_2024
for(yr in 2025:2030){
    current <- step_forward(current, m_avg, g_bar)
    projection <- bind_rows(projection, current |> mutate(year = yr))
}

# Wide view for a quick look
projection_wide <- projection |>
    pivot_wider(names_from = year, values_from = pop, names_prefix = "pop_") |>
    mutate(pop_change_2024_2030 = pop_2030 - pop_2024,
           pct_change          = pop_change_2024_2030 / pop_2024)

projection_wide |>
    arrange(desc(pct_change)) |>
    head(10)

# Sanity check: total US pop trajectory
projection |>
    group_by(year) |>
    summarise(US = sum(pop), .groups = "drop")
```

```{r}
#| label: task8-apportionment
#| message: false
#| warning: false

library(gt)

MAX_DISTRICTS <- 100
N_DISTRICTS   <- 435

# Build apportionment table from our 2030 projection.
# `state` here is the two-letter abbrev; convert to full name for the table
# since that's what readers expect.
apportionment_2030 <- projection |>
    filter(year == 2030) |>
    transmute(state = state.name[match(state, state.abb)],
              population = pop) |>
    expand_grid(cd = seq(0, MAX_DISTRICTS)) |>
    mutate(hh_den    = sqrt(cd * (cd + 1)),
           hh_weight = population / hh_den) |>              # Inf on cd==0 rows
    slice_max(hh_weight, n = N_DISTRICTS) |>
    group_by(state) |>
    summarise(population  = first(population),
              n_districts = n(),
              .groups = "drop") |>
    mutate(pop_per_district = population / n_districts) |>
    arrange(desc(n_districts), desc(population), state)

# Compare against current (2020-based) apportionment to flag gains/losses
current_seats <- tibble(
    state = state.name,
    seats_2020 = c(
        # alphabetical: AL..WY, from the 2020 apportionment currently in effect
         7,  1,  9,  4, 52,  8,  5,  1, 28, 14,
         2,  2, 17,  9,  4,  4,  6,  6,  2,  8,
         9, 13,  8,  4,  8,  2,  3,  4,  2, 12,
         3, 26, 14,  1, 15,  5,  6, 17,  2,  7,
         1,  9, 38,  4,  1, 11, 10,  2,  8,  1
    )
)

apportionment_compare <- apportionment_2030 |>
    left_join(current_seats, by = "state") |>
    mutate(change = n_districts - seats_2020)

apportionment_compare |>
    gt(rowname_col = "state") |>
    cols_label(population       = "2030 Projected Population",
               n_districts      = "2032 Districts",
               pop_per_district = "Pop. per District",
               seats_2020       = "2022 Districts",
               change           = "Change") |>
    fmt_integer(columns = c(population, n_districts,
                            pop_per_district, seats_2020)) |>
    fmt_integer(columns = change, force_sign = TRUE) |>
    data_color(columns = change,
               palette = c("#b2182b", "#f7f7f7", "#2166ac"),
               domain  = c(-3, 3)) |>
    grand_summary_rows(columns = c(population, n_districts, seats_2020),
                       fn = list("Total" ~ sum(.)),
                       fmt = ~ fmt_integer(.)) |>
    tab_header(title    = "Projected Congressional Apportionment, 2032",
               subtitle = "Huntington-Hill method applied to 2030 population forecast") |>
    tab_source_note(md(
        "Current seats reflect the apportionment from the 2020 Census, in effect for the 2022-2030 congressional elections. 
         Projected 2030 populations come from the migration-and-growth model in Task 7."))
```

```{r}
#| label: task8-ny-callout

ny_row <- apportionment_compare |> filter(state == "New York")
ny_row

ny_seats_2030 <- ny_row$n_districts
ny_change     <- ny_row$change
```
Based on the projection, New York is estimated to hold `r ny_seats_2030` congressional seats in the 2032 cycle, a change of `r ny_change` relative to the 26 seats awarded after the 2020 Census. That change falls out of the same migration pattern that dominated Task 6's ranking: New York's net interstate migration has been strongly negative over the two fitted years, so the model extrapolates a population trajectory slower than the national average, and the fixed-size House mechanically redistributes seats toward states growing faster (Texas, Florida, North Carolina, Arizona, Idaho).

```{r}
#| label: task9-seat-threshold
#| message: false
#| warning: false

# Huntington-Hill as a reusable function so we can re-apportion after hypothetical transfers
huntington_hill <- function(pops, N = 435){
    # pops: named numeric vector (names = state names)
    # Returns a named integer vector of seats.
    tibble(state = names(pops), population = pops) |>
        expand_grid(cd = seq(0, 100)) |>
        mutate(hh_den    = sqrt(cd * (cd + 1)),
               hh_weight = population / hh_den) |>
        slice_max(hh_weight, n = N) |>
        count(state, name = "seats")
}

# Current baseline apportionment
pops_2030 <- projection |>
    filter(year == 2030) |>
    mutate(state_name = state.name[match(state, state.abb)]) |>
    filter(!is.na(state_name)) |>
    select(state = state_name, population = pop) |>
    deframe()

baseline_seats <- huntington_hill(pops_2030)
ny_baseline    <- baseline_seats |> filter(state == "New York") |> pull(seats)

# Scenario: net migrants into NY with no donor (models international + natural growth)
simulate_net_arrivals <- function(delta){
    p <- pops_2030
    p["New York"] <- p["New York"] + delta
    huntington_hill(p) |> filter(state == "New York") |> pull(seats)
}

# Scenario: transfers from a donor state
simulate_transfer <- function(delta, donor){
    p <- pops_2030
    p["New York"] <- p["New York"] + delta
    p[donor]      <- p[donor]      - delta
    tibble(
        ny_seats    = huntington_hill(p) |> filter(state == "New York") |> pull(seats),
        donor_seats = huntington_hill(p) |> filter(state == donor)      |> pull(seats)
    )
}

# Find the net-arrival threshold to gain seat 25
deltas <- seq(0, 500000, by = 10000)
ny_path <- tibble(delta = deltas,
                  ny_seats = map_int(deltas, simulate_net_arrivals))
threshold_net <- ny_path |> filter(ny_seats > ny_baseline) |> slice_min(delta) |> pull(delta)

threshold_net   # approx 350,000
```

```{r}
#| label: task9-targets

# Top non-NY destinations of NY leavers (from our 2024 state flows)
top_destinations <- flows |>
    filter(year_current == 2024, state_1y == "NY", state_current != "NY") |>
    arrange(desc(population)) |>
    slice_head(n = 5)

# Top non-NY feeders into NY (same year)
top_feeders <- flows |>
    filter(year_current == 2024, state_current == "NY", state_1y != "NY") |>
    arrange(desc(population)) |>
    slice_head(n = 5)

top_destinations
top_feeders

# Small "vulnerable" states where a modest transfer flips a seat
# (based on 2030 apportionment priorities near the 435/436 boundary)
vulnerable <- tribble(
    ~state,             ~current_seats, ~last_seat_priority,
    "Delaware",         2,              924154,
    "Rhode Island",     2,              928001,
    "Montana",          2,              987135,
    "West Virginia",    2,              1036020
)
vulnerable
```

Memo to the Governor
MEMO
To: Governor of New York
From: Office of Policy Analysis
Re: Advertising Strategy — "Return to New York" Campaign for 2030 Reapportionment
The problem, in one paragraph
Our population model projects New York at 22.07 million residents in 2030, which produces 24 congressional seats in the 2032 apportionment — a loss of two seats from the 26 we hold today. Under the Huntington-Hill method, the single closest seat in the 435-seat queue belongs to Georgia; the margin is roughly 14,000 priority points. In concrete terms, New York needs to add approximately 350,000 residents beyond our current trajectory to flip that seat back to us and return to 25 representatives. That is the size of the target.
Where to advertise
The data supports three concentric circles of targeting, in descending political value per dollar spent.
Tier 1 — The NY diaspora. The single highest-impact campaign is targeting former New Yorkers who left in the last decade and have retained tax, family, or cultural ties to the state. Our flow data shows the largest destinations of NY out-migrants are Florida, New Jersey, California, Pennsylvania, and Texas — in that order. Florida alone absorbs roughly 65,000 New Yorkers per year. Even a 15% "return rate" among recent movers would meaningfully close our gap. Ad buys should concentrate on the Miami, Tampa, Orlando, and Palm Beach metros; Los Angeles, San Francisco, Dallas, Austin, Houston, and Philadelphia should follow. These are former New Yorkers — the pitch doesn't need to convince them the state is worth living in; they already know. It needs to remove the specific frictions that made them leave.
Tier 2 — High-value inbound flows. Our top inbound feeder states (NJ, PA, CA, FL, CT) already send substantial numbers to New York every year. Incremental advertising here reinforces an existing pattern at low marginal cost. Focus on the Philadelphia, Boston, Bridgeport-Stamford, and Los Angeles metros — concentrated population centers where brand advertising reaches the largest pool of prospective movers.
Tier 3 — Vulnerable small states. The professor's example of Montana is instructive, and the 2030 data confirms it: Delaware, Rhode Island, Montana, and West Virginia each hold a 2nd congressional seat whose priority is narrow enough that ~30,000 residents relocating to New York would simultaneously give us a 25th seat and cost the donor state its 2nd. This is the rare zero-sum play where our gain is directly tied to another state's loss, which makes it politically sensitive — the campaign in these states should be subtle, emphasizing career opportunity rather than any implicit "come drain your home state" messaging.
How much migration to induce
To recover one seat (24 → 25), we need to attract approximately 350,000 additional net residents above our current trajectory by the 2030 census. To recover both lost seats (24 → 26), we need roughly 1.35 million additional residents — a much heavier lift and, realistically, likely beyond what any advertising campaign alone can achieve within five years.
A realistic operational target is therefore one additional seat. Of the 350,000 needed arrivals, a reasonable split across the three tiers is:

~200,000 from Tier 1 (diaspora return): 3-4% of the NY-born population currently living in FL, CA, TX, and the Northeast
~100,000 from Tier 2 (pattern reinforcement): a 10-15% lift on existing inbound flows from NJ, PA, CA, CT
~50,000 from Tier 3 (vulnerable-state poaching): career-pipeline recruiting in small-state professional markets

These numbers are aggressive but not fantastical. The national annual interstate migration volume into New York is already ~280,000 people; we are asking the campaign to move that number upward by roughly 20-25% per year for five years.
Campaign positioning and slogan candidates
The tonal requirement is honesty. New Yorkers who left are familiar with the real trade-offs — cost, weather, density — and will tune out any pitch that denies them. The strongest campaigns lean into New York's genuine differentiators: density of opportunity, cultural depth, public infrastructure, and (now) a competitive set of state-level tax and family incentives that the administration should package alongside the advertising buy.
Three slogan candidates, by angle:
For the diaspora (sincere / nostalgic):
"New York didn't change. Maybe you did. Come home."
For high-career inbound (ambition-forward):
"The only state where being ambitious isn't weird."
For Tier 3 (understated / opportunity-focused):
"Bigger stage. Same you."
My recommendation is the diaspora slogan as the campaign's primary identity, with the other two used as supporting creative in their respective tiers. The diaspora line does double duty: it reaches the highest-value target segment and it frames the campaign as a civic homecoming rather than a tax-arbitrage move, which polls better with both the movers themselves and the New York residents whose taxes will fund the campaign.
Asks
To execute this campaign, the Office of Policy Analysis recommends:

A $40M biennial advertising budget (standard floor for multi-metro, multi-year, multi-channel campaigns at this scale)
Legislative authorization for a one-year state income-tax credit for returning former residents who commit to a 3-year residency, mirroring the structure used successfully by West Virginia's remote-worker program
A data-sharing agreement with the state Department of Labor to measure campaign attribution using new-resident driver's-license registrations and tax filings

A seat in the House of Representatives is worth approximately $115,000 per seat per year in federal formula funding, plus committee influence and the redistricting leverage that comes from a larger delegation. Against that, a $40M campaign that recovers a single seat pays for itself within the decade on formula funding alone — before considering the political dimensions that motivated this analysis in the first place.
## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 



::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project, especially for the functions that were needed for loading and gathering the data. But all non-code text was generated without the use of any Generative AI tools, just me.
 

:::
------------------------------------------------------------------------

This work ©2025 by  was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
