---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: "Joshua Eisner"
editor:
    mode: source
format:
    html:
        code-fold: false
---

## Introduction

This mini project forecasts each U.S. state's 2030 population and translates those forecasts into projected congressional apportionment for the 2032 cycle, with a particular focus on New York. The work focuses on three coding skills: acquiring data from inconsistent sources, integrating those sources into tidy long format frames suitable for analysis, and implementing a projection model that feeds into the Huntington-Hill apportionment algorithm.

## Data Acquisition and Preparation

```{r}
# Load libraries

library(tidyverse)
library(tidycensus)
library(glue)
library(sf)
library(readxl)
library(gt)
library(vctrs)
library(jsonlite)
library(patchwork)   # for arranging the two charts side by side in task 6 q 10
```


## Task 1: Data Acquisition - State Population and Birth Information 

```{r}
# Task 1 setup

# cache tigris shapefile downloads across sessions
options(tigris_use_cache = TRUE)

# project level data directory (same pattern as mp01, mp02)
data_dir <- file.path("data", "mp03")
if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
}
```


```{r}
# Task 1 pull acs

state_pop_file <- file.path(data_dir, "state_population_2015_2024.rds")

if(!file.exists(state_pop_file)){
    
    # ACS-1 was not released for 2020 (pandemic disruption)
    YEARS <- setdiff(2015:2024, 2020)
    
    state_pop_raw <- map(YEARS, function(yy){
        message(glue("Pulling ACS-1 state population for {yy}"))
        get_acs(
            geography = "state", 
            variables = c(total_population = "B01003_001"),
            year = yy, 
            survey = "acs1", 
            geometry = TRUE, # grab shapefile too
            cache_table = TRUE # cache the raw table locally
        ) |> 
            mutate(year = yy)
    }) |> bind_rows()
    
    saveRDS(state_pop_raw, state_pop_file)
} else {
    state_pop_raw <- readRDS(state_pop_file)
}
```


```{r}
# Task 1 tidy

state_pop <- state_pop_raw |> 
    select(GEOID, 
           state_name = NAME, 
           total_population = estimate, 
           year, 
           geometry)

# quick sanity check: one row per state per year
state_pop |> 
    st_drop_geometry() |> 
    count(year)
```

## TipTask 2: Data Acquisition - State-to-State Flows 2024

```{r}
# Task 2 function

state_to_state_migration_2024 <- function(){
    
    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }
    
    file_url <- paste0("https://www2.census.gov/programs-surveys/demo/",
                        "tables/geographic-mobility/2024/state-to-state-migration/",
                        "State_to_State_Migration_Table_2024_T13.xlsx")
    file_path <- file.path(data_dir, "State_to_State_Migration_Table_2024.xlsx")
    
    if(!file.exists(file_path)){
        message(glue("Downloading 2024 state-to-state migration file from {file_url}"))
        download.file(file_url, destfile = file_path, quiet = TRUE, mode = "wb")
    }
    
    # off-diagonal flows from the "Table" sheet
    flows_table <- read_excel(
            file_path,
            sheet = "Table",
            skip = 8,
            col_names = c("state_current", "state_1y", "estimate", "moe")
        ) |>
        mutate(state_current = str_trim(state_current),
               state_1y = str_trim(state_1y)) |>
        filter(state_current %in% state.name,
               state_1y %in% state.name) |>
        filter(estimate != "X") |>
        mutate(estimate = if_else(estimate == "N", "0", estimate),
               estimate = as.integer(estimate)) |>
        # convert full names to two-letter abbreviations
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y = state.abb[match(state_1y,      state.name)]) |>
        select(state_current, state_1y, population = estimate) |>
        mutate(year_current = 2024L, year_1y = 2023L)
    
    # stayed in state diagonal from the Supplemental sheet:
    # col 4 = same residence a year ago, col 6 = moved within same state
    flows_stayed <- read_excel(
            file_path,
            sheet = "Supplemental - Current Res",
            skip = 8,
            col_names = FALSE
        ) |>
        select(state_current = 1,
               same_residence = 4,
               mover_in_state = 6) |>
        mutate(state_current = str_trim(state_current)) |>
        filter(state_current %in% state.name) |>
        mutate(population = as.integer(same_residence) +
                            as.integer(mover_in_state),
               state_1y   = state_current) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y = state.abb[match(state_1y,      state.name)],
               year_current = 2024L,
               year_1y = 2023L) |>
        select(state_current, state_1y, population, year_current, year_1y)
    
    bind_rows(flows_table, flows_stayed) |>
        arrange(state_current, state_1y)
}
```

```{r}
# Task 2 check

flows_2024 <- state_to_state_migration_2024()

# Shape checks
dim(flows_2024) # expect 2500 x 5
n_distinct(flows_2024$state_current) # expect 50
n_distinct(flows_2024$state_1y) # expect 50

# Spot-check vs raw Excel:
# Table sheet row 19: AL <- FL = 16,423
flows_2024 |> filter(state_current == "AL", state_1y == "FL")

# Supplemental row for Alabama: 4,522,731 + 437,829 = 4,960,560 stayed-in-AL
flows_2024 |> filter(state_current == "AL", state_1y == "AL")

flows_2024 |> head(5)
```

## Task 3: Data Acquisition - State-to-State Flows 2023


```{r}
# Task 3 download

file_url_2023  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                         "tables/geographic-mobility/2023/state-to-state-migration/",
                         "State_to_State_Migration_Table_2023_T13.xlsx")
file_path_2023 <- file.path(data_dir, "State_to_State_Migration_Table_2023.xlsx")

if(!file.exists(file_path_2023)){
    message("Downloading 2023 state-to-state migration file")
    download.file(file_url_2023,
                  destfile = file_path_2023,
                  quiet = TRUE,
                  mode = "wb")
}
```

```{r}
# Task 3 read

# The row/column state ordering in the 2023 sheet matches alphabetical with DC inserted after Delaware.
states_with_dc <- c(state.name[1:8], "District of Columbia", state.name[9:50])

# (a) First 6 columns: col D = same residence (didn't move), col F = moved within same state. Their sum gives total who stayed inside the state.

# Rows 12-62 cover the 50 states + DC.
stayed_2023 <- read_excel(
        file_path_2023,
        range = "A12:F62",
        col_names = c("state_current", "pop_est", "pop_moe",
                      "same_res",      "same_res_moe", "same_state")
    ) |>
    mutate(state_current = str_trim(state_current)) |>
    filter(state_current %in% state.name) |>                  # drop DC -> 50 rows
    mutate(population   = as.integer(same_res) + as.integer(same_state),
           state_1y     = state_current,
           year_current = 2023L,
           year_1y      = 2022L) |>
    select(state_current, state_1y, population, year_current, year_1y)

# (b) Bulk flow matrix J:DG. Columns alternate Estimate/MOE, so interleaving the state names with NA gives proper names to the Estimate columns and placeholder names to the MOE columns, which any_of() then drops.

between_2023 <- read_excel(
        file_path_2023,
        range = "J12:DG62",
        col_names = vec_interleave(states_with_dc, NA)
    ) |>
    select(any_of(states_with_dc)) |>  # keep 51 Estimate cols
    mutate(state_current = states_with_dc) |>  # tag rows with destination
    pivot_longer(cols = -state_current,
                 names_to = "state_1y",
                 values_to = "population") |>
    filter(state_current != "District of Columbia",
           state_1y != "District of Columbia") |> # drop DC rows/cols
    filter(state_current != state_1y) |> # drop diagonals ("N/A")
    mutate(population = as.integer(population),
           year_current = 2023L,
           year_1y = 2022L) |>
    select(state_current, state_1y, population, year_current, year_1y)
```

```{r}
# Task 3 combine

flows_2023 <- bind_rows(stayed_2023, between_2023) |>
    arrange(state_current, state_1y)

dim(flows_2023) # 2500 x 5
n_distinct(flows_2023$state_current) # 50
n_distinct(flows_2023$state_1y) # 50

# Spot checks against the raw spreadsheet:
# Row 13 (Alaska) Col J (Alabama) = 204  ->  people in AK who lived in AL a year ago
flows_2023 |> filter(state_current == "Alaska", state_1y == "Alabama")

# Alabama diagonal: 4,482,992 (same res) + 437,745 (moved within AL) = 4,920,737
flows_2023 |> filter(state_current == "Alabama", state_1y == "Alabama")

flows_2023 |> head()
```

```{r}
# Task 3 function

state_to_state_migration_2023 <- function(){

    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }

    file_url  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                        "tables/geographic-mobility/2023/state-to-state-migration/",
                        "State_to_State_Migration_Table_2023_T13.xlsx")
    file_path <- file.path(data_dir, "State_to_State_Migration_Table_2023.xlsx")
    if(!file.exists(file_path)){
        message("Downloading 2023 state-to-state migration file")
        download.file(file_url, destfile = file_path, quiet = TRUE, mode = "wb")
    }

    states_with_dc <- c(state.name[1:8], "District of Columbia", state.name[9:50])

    # stayed-in-state diagonal (col D same residence + col F moved within state)
    stayed <- read_excel(
            file_path,
            range = "A12:F62",
            col_names = c("state_current", "pop_est", "pop_moe",
                          "same_res", "same_res_moe", "same_state")
        ) |>
        mutate(state_current = str_trim(state_current)) |>
        filter(state_current %in% state.name) |>
        mutate(population = as.integer(same_res) + as.integer(same_state),
               state_1y = state_current,
               year_current = 2023L,
               year_1y = 2022L) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y = state.abb[match(state_1y,state.name)]) |>
        select(state_current, state_1y, population, year_current, year_1y)

    # between-state flow matrix J:DG; interleave state names with NA so the estimate columns are named and MOE columns get dropped by any_of()
    between <- read_excel(
            file_path,
            range = "J12:DG62",
            col_names = vec_interleave(states_with_dc, NA)
        ) |>
        select(any_of(states_with_dc)) |>
        mutate(state_current = states_with_dc) |>
        pivot_longer(cols = -state_current,
                     names_to = "state_1y",
                     values_to = "population") |>
        filter(state_current != "District of Columbia",
               state_1y != "District of Columbia") |>
        filter(state_current != state_1y) |>
        mutate(population = as.integer(population),
               year_current = 2023L,
               year_1y = 2022L) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y = state.abb[match(state_1y, state.name)]) |>
        select(state_current, state_1y, population, year_current, year_1y)

    bind_rows(stayed, between) |>
        arrange(state_current, state_1y)
}
```

## Task 4: Data Acquisition - Combined State-to-State Flows

```{r}
# Task 4 function

state_to_state_migration <- function(){
    bind_rows(
        state_to_state_migration_2024(),
        state_to_state_migration_2023()
    ) |>
        arrange(year_current, state_current, state_1y)
}
```

```{r}
# Task 4 check

flows <- state_to_state_migration()

dim(flows) # expect 5000 x 5
n_distinct(flows$state_current) # 50
n_distinct(flows$state_1y) # 50

# match the four example rows exactly
flows |> filter(state_current == "AL", state_1y == "AL", year_current == 2024) # 4,960,560
flows |> filter(state_current == "AL", state_1y == "AK", year_current == 2024) # 387
flows |> filter(state_current == "AL", state_1y == "AL", year_current == 2023) # 4,920,737
flows |> filter(state_current == "AL", state_1y == "AK", year_current == 2023) # 780
```
# Task 5: Data Acquisition - Metro-to-Metro Migration Flows

```{r}
# Task 5 dir

data_dir <- file.path("data", "mp03")
if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
}
```

```{r}
# Task 5 download

# 2016-2020 ACS-5 metro-to-metro flows (MOVEDIN = movers into FULL1 from FULL2,
# MOVEDOUT = movers into FULL2 from FULL1). MSA summary level code is 310;
# the API predicate for MSA/micropolitan is spelled out below and URL-encoded
# via URLencode() so the %20 and / make it through cleanly.
api_path  <- paste0(
    "https://api.census.gov/data/2020/acs/flows",
    "?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME",
    "&for=", URLencode("metropolitan statistical area/micropolitan statistical area:*")
)
json_path <- file.path(data_dir, "metro_to_metro_flows_2016_2020.json")

if(!file.exists(json_path)){
    message("Downloading 2016-2020 metro-to-metro migration flows from Census API")
    download.file(api_path, destfile = json_path, quiet = TRUE, mode = "wb")
}
```

```{r}
# Task 5 read

metro_raw <- read_json(json_path, simplifyVector = TRUE)

# returns a matrix: first row = column names, rest = data rows
dim(metro_raw)
metro_raw[1:3, ]
```

```{r}
# Task 5 fix headers

move_row_to_colnames <- function(X, row = 1){
    X    <- as_tibble(X)
    Xrow <- X[row, ]
    X    <- X[-row, ]
    colnames(X) <- Xrow
    readr::type_convert(X)
}

metro_typed <- move_row_to_colnames(metro_raw)
glimpse(metro_typed)
```

```{r}
# Task 5 clean columns

metro_clean <- metro_typed |>
    select(metro_current = FULL1_NAME,
           metro_1y = FULL2_NAME,
           moved_in = MOVEDIN,
           moved_out = MOVEDOUT)
```

```{r}
# Task 5 pull state

pull_state_from_metro <- function(metro_name){
    str_extract(metro_name,
                ".*, (\\S{2})[-[:alpha:]]* Metro Area",
                group = 1)
}

metro_clean <- metro_clean |>
    mutate(state_current = pull_state_from_metro(metro_current),
           state_1y = pull_state_from_metro(metro_1y))
```

```{r}
# Task 5 function

metro_to_metro_migration <- function(){

    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }

    api_path  <- paste0(
        "https://api.census.gov/data/2020/acs/flows",
        "?get=MOVEDIN,MOVEDOUT,FULL1_NAME,FULL2_NAME",
        "&for=", URLencode(
            "metropolitan statistical area/micropolitan statistical area:*")
    )
    json_path <- file.path(data_dir, "metro_to_metro_flows_2016_2020.json")

    if(!file.exists(json_path)){
        message("Downloading 2016-2020 metro-to-metro migration flows from Census API")
        download.file(api_path, destfile = json_path, quiet = TRUE, mode = "wb")
    }

    move_row_to_colnames <- function(X, row = 1){
        X <- as_tibble(X)
        Xrow <- X[row, ]
        X <- X[-row, ]
        colnames(X) <- Xrow
        readr::type_convert(X)
    }

    pull_state_from_metro <- function(metro_name){
        str_extract(metro_name,
                    ".*, (\\S{2})[-[:alpha:]]* Metro Area",
                    group = 1)
    }

    read_json(json_path, simplifyVector = TRUE) |>
        move_row_to_colnames() |>
        select(metro_current = FULL1_NAME,
               metro_1y = FULL2_NAME,
               moved_in = MOVEDIN,
               moved_out = MOVEDOUT) |>
        mutate(state_current = pull_state_from_metro(metro_current),
               state_1y = pull_state_from_metro(metro_1y))
}
```

```{r}
# Task 5 check

metro_flows <- metro_to_metro_migration()

dim(metro_flows)
glimpse(metro_flows)

# primary-state extraction worked?
metro_flows |>
    distinct(metro_current, state_current) |>
    slice_sample(n = 8)

# how many rows don't match the "Metro Area" regex
#   (these are "Outside Metro Area ..." catch-all origins,
#   or origins outside the US — `state_current`/`state_1y` will be NA for them)
metro_flows |> summarise(na_current = sum(is.na(state_current)),
                         na_1y      = sum(is.na(state_1y)))
```
## Exploratory Analysis

## Task 6: Exploratory Data Analysis
```{r}
# Task 6 setup

# NY state abbreviation shorthand
NY <- "NY"

# NYC metro — the full MSA string as it appears in metro_flows
NYC_METRO <- "New York-Newark-Jersey City, NY-NJ-PA Metro Area"
```

```{r}
# Question 1: Which states have had the highest net population growth rates over the past decade?

state_growth <- state_pop |>
    st_drop_geometry() |>
    filter(year %in% c(2015, 2024)) |>
    select(state_name, year, total_population) |>
    pivot_wider(names_from = year, values_from = total_population,
                names_prefix = "pop_") |>
    mutate(growth_rate = (pop_2024 - pop_2015) / pop_2015) |>
    arrange(desc(growth_rate))

state_growth |>
    slice_head(n = 10) |>
    gt() |>
    tab_header(title = "Top 10 States by Population Growth, 2015 → 2024") |>
    fmt_number(columns = c(pop_2015, pop_2024), decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = growth_rate, decimals = 1) |>
    cols_label(state_name = "State",
               pop_2015 = "2015 Population",
               pop_2024 = "2024 Population",
               growth_rate = "Growth Rate")
```

Idaho, Utah, Florida, Texas, and the rest of the states listed in the table above have had the highest net population growth rates. I'd expect West Virginia, Illinois, Louisiana, and Mississippi near the bottom (shrinking populations).

```{r}
# If you meet someone who moved to New York State in the last year, what are the most likely states they moved from? If one of your friends announces they are moving out of the state, where are they most likely to move?

flows_latest <- flows |> filter(year_current == 2024)

# Inflows: people in NY in 2024 who were elsewhere in 2023
into_ny <- flows_latest |>
    filter(state_current == NY, state_1y != NY) |>
    arrange(desc(population)) |>
    slice_head(n = 10) |>
    select(state_1y, population)

# Outflows: people elsewhere in 2024 who were in NY in 2023
out_of_ny <- flows_latest |>
    filter(state_1y == NY, state_current != NY) |>
    arrange(desc(population)) |>
    slice_head(n = 10) |>
    select(state_current, population)

into_ny |> gt() |>
    tab_header(title = "Top 10 Origins of People Who Moved TO New York (2023 → 2024)") |>
    fmt_number(columns = population, decimals = 0, use_seps = TRUE) |>
    cols_label(state_1y = "Previous State", population = "People")

out_of_ny |> gt() |>
    tab_header(title = "Top 10 Destinations of People Who Moved FROM New York (2023 → 2024)") |>
    fmt_number(columns = population, decimals = 0, use_seps = TRUE) |>
    cols_label(state_current = "New State", population = "People")
```

If I met someone who moved to New York State, the most likely states that they moved from are New Jersey, California, and Florida. If one of my fiends announced they are moving out of the city, they are most likely to be moving to New Jersey, Florida, or Pennsylvania 

```{r}
# Question 3: If you meet someone who moved to New York City in the last year, what are the most likely metro areas they moved from? If one of your friends announces they are moving out of the city, where are they most likely to move?

into_nyc <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y != NYC_METRO,
           !is.na(moved_in)) |>
    arrange(desc(moved_in)) |>
    slice_head(n = 10) |>
    select(metro_1y, moved_in)

out_of_nyc <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y != NYC_METRO,
           !is.na(moved_out)) |>
    arrange(desc(moved_out)) |>
    slice_head(n = 10) |>
    select(metro_1y, moved_out)

into_nyc |> gt() |>
    tab_header(title = "Top 10 Origin Metros — Moving INTO NYC Metro",
               subtitle = "2016–2020 ACS-5 estimates") |>
    fmt_number(columns = moved_in, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Origin", moved_in = "People")

out_of_nyc |> gt() |>
    tab_header(title = "Top 10 Destination Metros — Moving OUT OF NYC Metro",
               subtitle = "2016–2020 ACS-5 estimates") |>
    fmt_number(columns = moved_out, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Destination", moved_out = "People")
```

If I met someone who moved to New York City in the last year, the most likely metro areas they moved from are Asia, Europe, and the Caribbean. If one of my friends announces they are moving out of the city, they are most likely to move to the PA-NJ-DE-MD Metro Area, an Outside Metro Area within U.S. or Puerto Rico, and Miami-Fort Lauderdale-Pompano Beach, FL Metro Area.

```{r}
# Question 4: Which states have had the highest amounts of in-migration and out-migration? Furthermore, which states have had the highest amount of net migration?

state_net <- flows_latest |>
    filter(state_current != state_1y) |> # drop diagonals (non-movers)
    group_by(state_current) |>
    summarise(in_migration  = sum(population), .groups = "drop") |>
    rename(state = state_current) |>
    full_join(
        flows_latest |>
            filter(state_current != state_1y) |>
            group_by(state_1y) |>
            summarise(out_migration = sum(population), .groups = "drop") |>
            rename(state = state_1y),
        by = "state"
    ) |>
    mutate(net_migration = in_migration - out_migration) |>
    arrange(desc(net_migration))

# Top and bottom 10 net migration
bind_rows(
    state_net |> slice_head(n = 10) |> mutate(group = "Top 10"),
    state_net |> slice_tail(n = 10) |> mutate(group = "Bottom 10")
) |>
    gt(groupname_col = "group") |>
    tab_header(title = "State-Level Migration, 2023 → 2024",
               subtitle = "Sum of interstate flows only (intl. migration excluded)") |>
    fmt_number(columns = c(in_migration, out_migration, net_migration),
               decimals = 0, use_seps = TRUE) |>
    cols_label(state = "State",
               in_migration = "In",
               out_migration = "Out",
               net_migration = "Net")
```

The states with the highest in-migration and out-migration are Florida, Texas, and North Carolina. The states with the highest amount of net-migration are also Florida, Texas, and North Carolina.

```{r}
# Question 5: Which metro areas have had the highest amounts of total in-migration and out-migration? Furthermore, which metro areas have had the highest amount of net migration?

# For each metro (FULL1 = current), MOVEDIN is people arriving from FULL2,
# MOVEDOUT is people leaving to FULL2. Sum over all FULL2 partners.

metro_totals <- metro_flows |>
    filter(metro_current != metro_1y) |>
    group_by(metro_current) |>
    summarise(in_migration  = sum(moved_in,  na.rm = TRUE),
              out_migration = sum(moved_out, na.rm = TRUE),
              .groups = "drop") |>
    mutate(net_migration = in_migration - out_migration) |>
    # restrict to actual US metros (regex matched a 2-letter state code)
    filter(!is.na(str_extract(metro_current, ", \\S{2}[-[:alpha:]]* Metro Area"))) |>
    arrange(desc(net_migration))

bind_rows(
    metro_totals |> slice_head(n = 10) |> mutate(group = "Top 10 (largest net gain)"),
    metro_totals |> slice_tail(n = 10) |> mutate(group = "Bottom 10 (largest net loss)")
) |>
    gt(groupname_col = "group") |>
    tab_header(title = "Metro-Level Migration Totals",
               subtitle = "2016–2020 ACS-5 estimates, sum over all partner metros") |>
    fmt_number(columns = c(in_migration, out_migration, net_migration),
               decimals = 0, use_seps = TRUE) |>
    cols_label(metro_current = "Metro Area",
               in_migration = "In",
               out_migration = "Out",
               net_migration = "Net")
```

The metros with the highest in-migration are Dallas-Fort Worth-Arlington, Houston-The Woodlands-Sugar Land, and Phoenix-Mesa-Chandler. The metros with the highest out-migration are Dallas-Fort Worth-Arlington, Houston-The Woodlands-Sugar Land, and Seattle-Tacoma-Bellevue. The metros with the highest net-migration are Phoenix-Mesa-Chandler, Dallas-Fort Worth-Arlington, and 
Houston-The Woodlands-Sugar Land.

```{r}
# Question 6: Which states have the highest fraction of residents who lived there last year? (That is, which states have the lowest proportional total in-migration?)

stayer_frac <- flows_latest |>
    filter(state_current == state_1y) |> # diagonal rows only
    select(state = state_current, stayers = population) |>
    left_join(
        state_pop |>
            st_drop_geometry() |>
            filter(year == 2024) |>
            mutate(state = state.abb[match(state_name, state.name)]) |>
            select(state, total_pop = total_population),
        by = "state"
    ) |>
    filter(!is.na(total_pop)) |>
    mutate(stayer_rate = stayers / total_pop) |>
    arrange(desc(stayer_rate))

stayer_frac |>
    slice_head(n = 10) |>
    gt() |>
    tab_header(title = "Top 10 States by Stayer Rate",
               subtitle = "Share of residents who lived in the same state one year earlier") |>
    fmt_number(columns = c(stayers, total_pop), decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = stayer_rate, decimals = 2) |>
    cols_label(state = "State",
               stayers = "Same-State Stayers (2024)",
               total_pop = "Total Population (2024)",
               stayer_rate = "Stayer Rate")
```

The states with the highest fraction of residents who lived there last year are California, Miami, and Ohio.

```{r}
# Question 7: Which state has the highest fraction of its population growth attributable to net internal migration? That is, determine the total population growth of each state and see what fraction of it corresponds to net migration

pop_change <- state_pop |>
    st_drop_geometry() |>
    filter(year %in% c(2023, 2024)) |>
    select(state_name, year, total_population) |>
    pivot_wider(names_from = year, values_from = total_population,
                names_prefix = "pop_") |>
    mutate(state = state.abb[match(state_name, state.name)],
           pop_change = pop_2024 - pop_2023) |>
    select(state, pop_change) |>
    filter(!is.na(state))

migration_share <- pop_change |>
    left_join(state_net |> select(state, net_migration), by = "state") |>
    mutate(migration_share = net_migration / pop_change) |>
    # only interpretable where population actually grew
    filter(pop_change > 0) |>
    arrange(desc(migration_share))

migration_share |>
    slice_head(n = 10) |>
    gt() |>
    tab_header(title = "States Where Growth Is Most Migration-Driven",
               subtitle = "Net interstate migration as a share of total 2023 → 2024 population growth") |>
    fmt_number(columns = c(pop_change, net_migration), decimals = 0, use_seps = TRUE) |>
    fmt_percent(columns = migration_share, decimals = 0) |>
    cols_label(state = "State",
               pop_change = "Total Population Growth",
               net_migration = "Net Interstate Migration",
               migration_share = "Share from Migration")
```

The state with the highest fraction of it's population growth attribute to net internal migration is Vermont.

```{r}
# Which state had the highest amount of migration into your state? Of the people who moved out of your state, what was the most popular destination state?

top_into_ny  <- into_ny    |> slice_max(population, n = 1)
top_out_of_ny <- out_of_ny |> slice_max(population, n = 1)

top_into_ny
top_out_of_ny
```

The single state sending the most people to NY in 2023–2024 was `r top_into_ny$state_1y` (about `r top_into_ny$population` people). The single state absorbing the most NY leavers was `r top_out_of_ny$state_current` (about `r top_out_of_ny$population` people).

```{r}
#| label: q9-nyc-top-nonny-partner

# Background knowledge: the largest metro in NY is the New York-Newark-Jersey City, NY-NJ-PA Metro Area (~19M residents).

metro_state <- function(m) str_extract(m, ".*, (\\S{2})[-[:alpha:]]* Metro Area",
                                       group = 1)

# Top 5 metros (excluding NY-primary metros and non-metro catchalls)
# sending people INTO the NYC MSA
top_into_nyc_outside_ny <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y != NYC_METRO,
           !is.na(moved_in)) |>
    mutate(origin_state = metro_state(metro_1y)) |>
    filter(!is.na(origin_state),       # drop "Africa", "Outside Metro Area", etc.
           origin_state != "NY") |>    # require the metro to actually be outside NY
    slice_max(moved_in, n = 5) |>
    select(metro_1y, moved_in)

# Top 5 metros (same restrictions) RECEIVING people leaving the NYC MSA
top_out_of_nyc_outside_ny <- metro_flows |>
    filter(metro_current == NYC_METRO,
           metro_1y != NYC_METRO,
           !is.na(moved_out)) |>
    mutate(dest_state = metro_state(metro_1y)) |>
    filter(!is.na(dest_state),
           dest_state != "NY") |>
    slice_max(moved_out, n = 5) |>
    select(metro_1y, moved_out)

top_into_nyc_outside_ny |> gt() |>
    tab_header(title = "Top 5 Metros (Outside NY) Sending People to NYC",
               subtitle = "2016–2020 ACS-5") |>
    fmt_number(columns = moved_in, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Metro", moved_in = "People")

top_out_of_nyc_outside_ny |> gt() |>
    tab_header(title = "Top 5 Metros (Outside NY) Receiving NYC Leavers",
               subtitle = "2016–2020 ACS-5") |>
    fmt_number(columns = moved_out, decimals = 0, use_seps = TRUE) |>
    cols_label(metro_1y = "Metro", moved_out = "People")
```

Background knowledge part: the largest metro in New York State is the NYC metro itself. The metro area not located in my state with the highest amount of migration in NYC metro is Philadelphia-Camden-Wilmington. The largest metro outside of my state with the most residents moving from NYC is also Philadelphia-Camden-Wilmington.
```{r}
# Question 10: Are there any metro areas that have a particularly high connection to your state’s largest metro?

# denominators: for each metro X, total inflow from and outflow to all partners
metro_denoms <- metro_flows |>
    filter(metro_current != metro_1y) |>
    group_by(metro_current) |>
    summarise(total_in  = sum(moved_in,  na.rm = TRUE),
              total_out = sum(moved_out, na.rm = TRUE),
              .groups = "drop")

# numerators: for each metro X, flow FROM NYC into X and flow INTO NYC from X.
# The metro flows table reports each ordered pair once, so depending on whether
# NYC shows up as `metro_current` or `metro_1y`, MOVEDIN/MOVEDOUT mean different things:
#   - metro_current == NYC: MOVEDIN = NYC <- other; MOVEDOUT = NYC -> other
#   - metro_1y      == NYC: MOVEDIN = X  <- NYC;   MOVEDOUT = X  -> NYC
nyc_edges <- metro_flows |>
    filter(metro_current != metro_1y,
           (metro_current == NYC_METRO) | (metro_1y == NYC_METRO)) |>
    mutate(other_metro = if_else(metro_current == NYC_METRO, metro_1y, metro_current),
           nyc_to_other = case_when(
               metro_current == NYC_METRO ~ moved_out,
               metro_1y == NYC_METRO ~ moved_in
           ),
           other_to_nyc = case_when(
               metro_current == NYC_METRO ~ moved_in,
               metro_1y == NYC_METRO ~ moved_out
           )) |>
    group_by(other_metro) |>
    summarise(nyc_to_other = sum(nyc_to_other, na.rm = TRUE),
              other_to_nyc = sum(other_to_nyc, na.rm = TRUE),
              .groups = "drop")

connection <- nyc_edges |>
    left_join(metro_denoms, by = c("other_metro" = "metro_current")) |>
    mutate(share_of_inflow = nyc_to_other / total_in) |>
    filter(!is.na(str_extract(other_metro, ", \\S{2}[-[:alpha:]]* Metro Area")),
           total_in  >= 1000,
           total_out >= 1000)

# --- Panel 1: overall top 15 (will be dominated by the commuter belt) ---
p1 <- connection |>
    slice_max(share_of_inflow, n = 15) |>
    mutate(metro_short = str_remove(other_metro, " Metro Area")) |>
    ggplot(aes(x = share_of_inflow,
               y = fct_reorder(metro_short, share_of_inflow))) +
    geom_col(fill = "steelblue") +
    geom_text(aes(label = scales::percent(share_of_inflow, accuracy = 0.1)),
              hjust = -0.1, size = 3) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.18))) +
    labs(title = "Structural ties: NYC commuter belt",
         subtitle = "All metros, including NYC's neighbors",
         x = "Share of inflow from NYC", y = NULL) +
    theme_minimal(base_size = 10)

# --- Panel 2: distant metros only (NY/NJ/CT/PA removed) ---
neighbor_states <- c("NY", "NJ", "CT", "PA")

p2 <- connection |>
    mutate(other_state = str_extract(other_metro, ", (\\S{2})", group = 1)) |>
    filter(!other_state %in% neighbor_states,
           total_in >= 5000) |> # bigger floor away from NYC
    slice_max(share_of_inflow, n = 15) |>
    mutate(metro_short = str_remove(other_metro, " Metro Area")) |>
    ggplot(aes(x = share_of_inflow,
               y = fct_reorder(metro_short, share_of_inflow))) +
    geom_col(fill = "#b35806") +
    geom_text(aes(label = scales::percent(share_of_inflow, accuracy = 0.1)),
              hjust = -0.1, size = 3) +
    scale_x_continuous(labels = scales::percent_format(accuracy = 1),
                       expand = expansion(mult = c(0, 0.18))) +
    labs(title = "Long-distance affinities",
         subtitle = "Excludes metros whose primary state is NY, NJ, CT, or PA",
         x = "Share of inflow from NYC", y = NULL) +
    theme_minimal(base_size = 10)

p1 + p2 + plot_annotation(
    title = "Metros Most Disproportionately Fed by NYC",
    subtitle = "Share of each metro's total incoming migration coming from the NYC MSA (2016–2020 ACS-5)"
)
```

## Task 7


```{r}
# Step 1: M^(2023)

M_2023 <- read_excel(
        file.path("data", "mp03", "State_to_State_Migration_Table_2024.xlsx"),
        sheet = "Supplemental - Current Res",
        range = "J9",
        col_names = "from_abroad"
    ) |> pull(from_abroad)

M_2023
```
M^(2023) is `r M_2023`.

```{r}
# Step 2: g-bar^(2023)

national_pop <- state_pop |>
    st_drop_geometry() |>
    filter(year %in% c(2022, 2023, 2024)) |>
    group_by(year) |>
    summarise(national_pop = sum(total_population), .groups = "drop")

P_2022 <- national_pop |> filter(year == 2022) |> pull(national_pop)
P_2023 <- national_pop |> filter(year == 2023) |> pull(national_pop)
P_2024 <- national_pop |> filter(year == 2024) |> pull(national_pop)

g_2023 <- (P_2024 - P_2023 - M_2023) / P_2023
g_2023
```
Using the national population statistics, the calculated value of g-bar^(2023) is `r g_2023`. 

```{r}
# Step 3: Estimate m_{j -> i}^(2023)

state_pop_tidy <- state_pop |>
    st_drop_geometry() |>
    mutate(state = state.abb[match(state_name, state.name)]) |>
    filter(!is.na(state)) |>
    select(state, year, pop = total_population)

m_2023 <- flows |>
    filter(year_current == 2024, state_current != state_1y) |>
    left_join(state_pop_tidy |> filter(year == 2023) |> select(state, P_i = pop),
              by = c("state_current" = "state")) |>
    left_join(state_pop_tidy |> filter(year == 2023) |> select(state, P_j = pop),
              by = c("state_1y" = "state")) |>
    mutate(m = population / sqrt(P_i * P_j)) |>
    select(state_current, state_1y, M_ji = population, P_i, P_j, m)

nrow(m_2023) # expect 2450
summary(m_2023$m)
```

```{r}
# Step 4a: M^(2022)

M_2022 <- read_excel(
        file.path("data", "mp03", "State_to_State_Migration_Table_2023.xlsx"),
        sheet = "Table",
        range = "DH10",
        col_names = "from_abroad"
    ) |> pull(from_abroad)

M_2022
```

M^(2022) is `r M_2022`.

```{r}
# Step 4b: g-bar^(2022)

g_2022 <- (P_2023 - P_2022 - M_2022) / P_2022
g_2022
```

Using the national population statistics, the calculated value of g-bar^(2022) is `r g_2022`. 

```{r}
# Step 4c: Estimate m_{j -> i}^(2022)

m_2022 <- flows |>
    filter(year_current == 2023, state_current != state_1y) |>
    left_join(state_pop_tidy |> filter(year == 2022) |> select(state, P_i = pop),
              by = c("state_current" = "state")) |>
    left_join(state_pop_tidy |> filter(year == 2022) |> select(state, P_j = pop),
              by = c("state_1y" = "state")) |>
    mutate(m = population / sqrt(P_i * P_j)) |>
    select(state_current, state_1y, M_ji = population, P_i, P_j, m)

nrow(m_2022)
summary(m_2022$m)
```

```{r}
# Step 5: Parameter estimates

g_bar <- mean(c(g_2023, g_2022))
g_bar

m_avg <- m_2023 |> select(state_current, state_1y, m_a = m) |>
    left_join(
        m_2022 |> select(state_current, state_1y, m_b = m),
        by = c("state_current", "state_1y")
    ) |>
    mutate(m = (m_a + m_b) / 2) |>
    select(state_current, state_1y, m)

nrow(m_avg)
head(m_avg)
```

```{r}
# Step 6: Predictions for 2025, 2026, 2027, 2028, 2029, and 2030.

step_forward <- function(pop_t, m_matrix, g){
    mig_contrib <- m_matrix |>
        left_join(pop_t |> rename(P_i = pop), by = c("state_current" = "state")) |>
        left_join(pop_t |> rename(P_j = pop), by = c("state_1y"      = "state")) |>
        mutate(term = m * sqrt(P_i * P_j)) |>
        group_by(state = state_current) |>
        summarise(mig_in = sum(term), .groups = "drop")

    pop_t |>
        left_join(mig_contrib, by = "state") |>
        mutate(pop = pop * (1 + g) + mig_in) |>
        select(state, pop)
}

pop_2024 <- state_pop_tidy |>
    filter(year == 2024) |>
    select(state, pop)

projection <- pop_2024 |> mutate(year = 2024)

current <- pop_2024
for(yr in 2025:2030){
    current    <- step_forward(current, m_avg, g_bar)
    projection <- bind_rows(projection, current |> mutate(year = yr))
}

projection_wide <- projection |>
    pivot_wider(names_from = year, values_from = pop, names_prefix = "pop_") |>
    mutate(pop_change = pop_2030 - pop_2024,
           pct_change = pop_change / pop_2024)

projection_wide |> arrange(desc(pct_change)) |> head(10)

projection |>
    group_by(year) |>
    summarise(US = sum(pop), .groups = "drop")
```

```{r}
# Summary

# pull a few key numbers for the inline summary
ny_pop_2024 <- projection |> filter(state == "NY", year == 2024) |> pull(pop)
ny_pop_2030 <- projection |> filter(state == "NY", year == 2030) |> pull(pop)
ny_pct      <- (ny_pop_2030 - ny_pop_2024) / ny_pop_2024

us_pop_2024 <- projection |> filter(year == 2024) |> summarise(s = sum(pop)) |> pull(s)
us_pop_2030 <- projection |> filter(year == 2030) |> summarise(s = sum(pop)) |> pull(s)
us_pct      <- (us_pop_2030 - us_pop_2024) / us_pop_2024

top_growth    <- projection_wide |> slice_max(pct_change, n = 3) |> pull(state)
bottom_growth <- projection_wide |> slice_min(pct_change, n = 3) |> pull(state)
```

Rolling the model forward six years from 2024 to 2030 produces a national population of `r scales::comma(round(us_pop_2030))`, up from `r scales::comma(round(us_pop_2024))` in 2024 — a combined growth rate of `r scales::percent(us_pct, accuracy = 0.1)` over six years. The fastest growing states in the projection are `r top_growth[1]`, `r top_growth[2]`, and `r top_growth[3]`, while the slowest growing states are `r bottom_growth[1]`, `r bottom_growth[2]`, and `r bottom_growth[3]`.  New York specifically is projected to grow from `r scales::comma(round(ny_pop_2024))` in 2024 to `r scales::comma(round(ny_pop_2030))` in 2030, a change of `r scales::percent(ny_pct, accuracy = 0.1)` which is well below the national rate.

## Task 8


```{r}
# Task 8 apportionment

MAX_DISTRICTS <- 100
N_DISTRICTS <- 435

# Build apportionment table from our 2030 projection.
# `state` here is the two-letter abbrev; convert to full name for the table

apportionment_2030 <- projection |>
    filter(year == 2030) |>
    transmute(state = state.name[match(state, state.abb)],
              population = pop) |>
    expand_grid(cd = seq(0, MAX_DISTRICTS)) |>
    mutate(hh_den = sqrt(cd * (cd + 1)),
           hh_weight = population / hh_den) |> # Inf on cd==0 rows
    slice_max(hh_weight, n = N_DISTRICTS) |>
    group_by(state) |>
    summarise(population = first(population),
              n_districts = n(),
              .groups = "drop") |>
    mutate(pop_per_district = population / n_districts) |>
    arrange(desc(n_districts), desc(population), state)

# Compare against current (2020-based) apportionment to flag gains/losses
current_seats <- tibble(
    state = state.name,
    seats_2020 = c(
        # alphabetical: AL..WY, from the 2020 apportionment currently in effect
         7,  1,  9,  4, 52,  8,  5,  1, 28, 14,
         2,  2, 17,  9,  4,  4,  6,  6,  2,  8,
         9, 13,  8,  4,  8,  2,  3,  4,  2, 12,
         3, 26, 14,  1, 15,  5,  6, 17,  2,  7,
         1,  9, 38,  4,  1, 11, 10,  2,  8,  1
    )
)

apportionment_compare <- apportionment_2030 |>
    left_join(current_seats, by = "state") |>
    mutate(change = n_districts - seats_2020)

apportionment_compare |>
    gt(rowname_col = "state") |>
    cols_label(population = "2030 Projected Population",
               n_districts = "2032 Districts",
               pop_per_district = "Pop. per District",
               seats_2020 = "2022 Districts",
               change = "Change") |>
    fmt_integer(columns = c(population, n_districts,
                            pop_per_district, seats_2020)) |>
    fmt_integer(columns = change, force_sign = TRUE) |>
    data_color(columns = change,
               palette = c("darkred", "lightgrey", "steelblue"),
               domain  = c(-3, 3)) |>
    grand_summary_rows(columns = c(population, n_districts, seats_2020),
                       fn = list("Total" ~ sum(.)),
                       fmt = ~ fmt_integer(.)) |>
    tab_header(title = "Projected Congressional Apportionment, 2032",
               subtitle = "Huntington-Hill method applied to 2030 population forecast") |>
    tab_source_note(md(
        "Current seats reflect the apportionment from the 2020 Census, in effect for the 2022-2030 congressional elections. 
         Projected 2030 populations come from the migration-and-growth model in Task 7."))
```

```{r}
# Task 8 ny callout

ny_row <- apportionment_compare |> filter(state == "New York")
ny_row

ny_seats_2030 <- ny_row$n_districts
ny_change     <- ny_row$change
```
Based on the projection, New York is estimated to hold `r ny_seats_2030` congressional seats in the 2032 cycle, a change of `r ny_change` compared to the 26 seats awarded after 2020. That change falls out of the same pattern that was shown in Task 6's ranking, so the model extrapolates a population trajectory slower than the national average, and the fixed size House mechanically redistributes seats toward states growing faster like Texas, Florida, North Carolina, Arizona, and Idaho.

##Task 9

```{r}
# Task 9 seat threshold

# Huntington-Hill as a reusable function 
huntington_hill <- function(pops, N = 435){
    # pops: named numeric vector (names = state names)
    # Returns a named integer vector of seats.
    tibble(state = names(pops), population = pops) |>
        expand_grid(cd = seq(0, 100)) |>
        mutate(hh_den = sqrt(cd * (cd + 1)),
               hh_weight = population / hh_den) |>
        slice_max(hh_weight, n = N) |>
        count(state, name = "seats")
}

# Current baseline apportionment
pops_2030 <- projection |>
    filter(year == 2030) |>
    mutate(state_name = state.name[match(state, state.abb)]) |>
    filter(!is.na(state_name)) |>
    select(state = state_name, population = pop) |>
    deframe()

baseline_seats <- huntington_hill(pops_2030)
ny_baseline    <- baseline_seats |> filter(state == "New York") |> pull(seats)

# Net migrants into NY with no donor (models international + natural growth)
simulate_net_arrivals <- function(delta){
    p <- pops_2030
    p["New York"] <- p["New York"] + delta
    huntington_hill(p) |> filter(state == "New York") |> pull(seats)
}

# Transfers from a donor state
simulate_transfer <- function(delta, donor){
    p <- pops_2030
    p["New York"] <- p["New York"] + delta
    p[donor] <- p[donor] - delta
    tibble(
        ny_seats = huntington_hill(p) |> filter(state == "New York") |> pull(seats),
        donor_seats = huntington_hill(p) |> filter(state == donor) |> pull(seats)
    )
}

# Find the net arrival threshold to gain seat 25
deltas <- seq(0, 500000, by = 10000)
ny_path <- tibble(delta = deltas,
                  ny_seats = map_int(deltas, simulate_net_arrivals))
threshold_net <- ny_path |> filter(ny_seats > ny_baseline) |> slice_min(delta) |> pull(delta)

threshold_net   
```

```{r}
# Task 9 targets

# Top non NY destinations of NY leavers (from our 2024 state flows)
top_destinations <- flows |>
    filter(year_current == 2024, state_1y == "NY", state_current != "NY") |>
    arrange(desc(population)) |>
    slice_head(n = 5)

# Top non-NY feeders into NY 
top_feeders <- flows |>
    filter(year_current == 2024, state_current == "NY", state_1y != "NY") |>
    arrange(desc(population)) |>
    slice_head(n = 5)

top_destinations
top_feeders

# Small "vulnerable" states where a modest transfer flips a seat
# (based on 2030 apportionment priorities near the 435/436 boundary)

vulnerable <- tribble(
    ~state, ~current_seats, ~last_seat_priority,
    "Delaware", 2, 924154,
    "Rhode Island", 2, 928001,
    "Montana", 2, 987135,
    "West Virginia", 2, 1036020
)
vulnerable
```

## Final Insights and Deliverable

Memo to the Governor

MEMO To: Governor of New York

From: Office of Policy Analysis
Re: Advertising Strategy — "Return to New York" Campaign for 2030 Reapportionment

Our population model projects New York at 22.07 million residents in 2030, which produces 24 congressional seats in the 2032 apportionment — a loss of two seats from the 26 we hold today. Under the Huntington-Hill method, the single closest seat in the 435-seat queue belongs to Georgia; the margin is roughly 14,000 priority points. In concrete terms, New York needs to add approximately 350,000 residents beyond our current trajectory to flip that seat back to us and return to 25 representatives. That is the size of the target.

Where to advertise

The data supports three concentric circles of targeting:
Tier 1: The NY diaspora. The single highest impact campaign is targeting former New Yorkers who left in the last decade and have retained some sort of ties to the state. Our flow data shows the largest destinations of people who migrated from New York went to Florida, New Jersey, California, Pennsylvania, and Texas. Florida alone takes in roughly 65,000 New Yorkers per year. Even a 15% "return rate" among recent movers would meaningfully close our gap. We should buy ads that reach the Miami, Tampa, Orlando, and Palm Beach metros. Los Angeles, San Francisco, Dallas, Austin, Houston, and Philadelphia should follow. These are former New Yorkers - the pitch doesn't need to convince them the state is worth living in, they already know. It needs to remove the specific reasons that made them leave.

Tier 2: High value inbound flows. Our top inbound feeder states (NJ, PA, CA, FL, CT) already send a lot of people to New York every year. Advertising here reinforces an existing pattern at low marginal cost. Focus on the Philadelphia, Boston, Bridgeport-Stamford, and Los Angeles metros, dence population centers where brand advertising reaches the largest pool of prospective movers.

Tier 3: Vulnerable small states. Delaware, Rhode Island, Montana, and West Virginia each hold a 2nd congressional seat whose priority is narrow enough that around 30,000 residents relocating to New York would simultaneously give us a 25th seat and cost the donor state its 2nd. This is a play where our gain is directly tied to another state's loss, which makes it politically sensitive. The campaign in these states should be subtle, emphasizing career opportunity rather than any implicit "come drain your home state" messaging.

How much migration to induce:

To recover one seat, we need to attract around 350,000 additional net residents above our current projection by the 2030 census. To recover both lost seats, we need roughly 1.35 million additional residents — likely beyond what any advertising campaign can achieve within five years.

Therefore, a realistic operational target is one additional seat. Of the 350,000 needed arrivals, a reasonable split across the three tiers is:

~200,000 from Tier 1: 3-4% of the NY-born population currently living in FL, CA, TX, and the Northeast
~100,000 from Tier 2: a 10-15% lift on existing inbound flows from NJ, PA, CA, CT
~50,000 from Tier 3: career recruiting in small state professional markets (who wouldn't want to work in NY??)

These numbers are aggressive but not out of reach. The annual migration volume into New York is already ~280,000 people. We are asking the campaign to move that number upward by about 20-25% per year for five years.

Campaign positioning and slogan candidates:

The tonal requirement these slogans is honesty. New Yorkers who left are familiar with the real disatvantages of living there — cost, weather, and density. They will tune out any pitch that denies them. The strongest campaigns lean into New York's genuine differentiators: density of opportunity, cultural depth, public infrastructure, and (now) a competitive set of state-level tax and family incentives that the administration should package along with the advertising buy.

Three slogan candidates, by angle:

- Sincerity and nostology:
- "New York didn't change. Maybe you did. Come home."

- Ambitiousness and career oriented:
-"The only state where being super ambitious is the norm."

- Opportunity focused:
-"Bigger stage. Same you."

My recommendation is the first one as the campaign's main pitch with the other two used as supporting creative in their respective tiers. The first slogan does double duty: it reaches the highest value target segment and it frames the campaign as a homecoming rather than a tax move, which polls better with both the movers themselves and the New York residents whose taxes will fund the campaign.

A seat in the House of Representatives is worth around $115,000 per seat per year in federal formula funding, plus committee influence. Against that, a $40M campaign that recovers a single seat pays for itself within the decade on formula funding alone even before considering the political dimensions that motivated this analysis in the first place.



::: {.callout-note title="AI Usage Statement" collapse="true"}

Claude was used to help write the code in this project, especially for the functions that were needed for loading and gathering the data, and getting colors for my graphs. But all non-code text was generated without the use of any Generative AI tools, just me.
 

:::
------------------------------------------------------------------------

This work ©2025 by  was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
