**This is only a basic Mini-Project Template.**
**Delete this note and customize formatting according to your preferences and**
**the Mini-Project instructions. Your document must start the `---` below**
---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: ""
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

Remember that code blocks look like this: 

```{r}
#| label: task1-setup
#| message: false
#| warning: false

library(tidyverse)
library(tidycensus)
library(glue)
library(sf)
library(readxl)

# cache tigris shapefile downloads across sessions
options(tigris_use_cache = TRUE)

# project-level data directory (same pattern as mp01, mp02)
data_dir <- file.path("data", "mp03")
if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
}
```


```{r}
#| label: task1-pull-acs1
#| message: false

state_pop_file <- file.path(data_dir, "state_population_2015_2024.rds")

if(!file.exists(state_pop_file)){
    
    # ACS-1 was not released for 2020 (pandemic disruption)
    YEARS <- setdiff(2015:2024, 2020)
    
    state_pop_raw <- map(YEARS, function(yy){
        message(glue("Pulling ACS-1 state population for {yy}"))
        get_acs(
            geography = "state", 
            variables = c(total_population = "B01003_001"),
            year      = yy, 
            survey    = "acs1", 
            geometry  = TRUE,    # grab shapefile too
            cache_table = TRUE   # cache the raw table locally
        ) |> 
            mutate(year = yy)
    }) |> bind_rows()
    
    saveRDS(state_pop_raw, state_pop_file)
} else {
    state_pop_raw <- readRDS(state_pop_file)
}
```


```{r}
#| label: task1-tidy

state_pop <- state_pop_raw |> 
    select(GEOID, 
           state_name       = NAME, 
           total_population = estimate, 
           year, 
           geometry)

# quick sanity check: one row per state per year
state_pop |> 
    st_drop_geometry() |> 
    count(year)
```

```{r}
#| label: task2-function
#| message: false
#| warning: false

library(tidyverse)
library(readxl)
library(glue)

state_to_state_migration_2024 <- function(){
    
    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }
    
    file_url  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                        "tables/geographic-mobility/2024/state-to-state-migration/",
                        "State_to_State_Migration_Table_2024_T13.xlsx")
    file_path <- file.path(data_dir, "State_to_State_Migration_Table_2024.xlsx")
    
    if(!file.exists(file_path)){
        message(glue("Downloading 2024 state-to-state migration file from {file_url}"))
        download.file(file_url, destfile = file_path, quiet = TRUE, mode = "wb")
    }
    
    # off-diagonal flows from the "Table" sheet
    flows_table <- read_excel(
            file_path,
            sheet     = "Table",
            skip      = 8,
            col_names = c("state_current", "state_1y", "estimate", "moe")
        ) |>
        mutate(state_current = str_trim(state_current),
               state_1y      = str_trim(state_1y)) |>
        filter(state_current %in% state.name,
               state_1y      %in% state.name) |>
        filter(estimate != "X") |>
        mutate(estimate = if_else(estimate == "N", "0", estimate),
               estimate = as.integer(estimate)) |>
        # convert full names to two-letter abbreviations
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)]) |>
        select(state_current, state_1y, population = estimate) |>
        mutate(year_current = 2024L, year_1y = 2023L)
    
    # stayed-in-state diagonal from the Supplemental sheet:
    # col 4 = same residence a year ago, col 6 = moved within same state
    flows_stayed <- read_excel(
            file_path,
            sheet     = "Supplemental - Current Res",
            skip      = 8,
            col_names = FALSE
        ) |>
        select(state_current  = 1,
               same_residence = 4,
               mover_in_state = 6) |>
        mutate(state_current = str_trim(state_current)) |>
        filter(state_current %in% state.name) |>
        mutate(population = as.integer(same_residence) +
                            as.integer(mover_in_state),
               state_1y   = state_current) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)],
               year_current  = 2024L,
               year_1y       = 2023L) |>
        select(state_current, state_1y, population, year_current, year_1y)
    
    bind_rows(flows_table, flows_stayed) |>
        arrange(state_current, state_1y)
}
```

```{r}
#| label: task2-check

flows_2024 <- state_to_state_migration_2024()

# Shape checks
dim(flows_2024)                         # expect 2500 x 5
n_distinct(flows_2024$state_current)    # expect 50
n_distinct(flows_2024$state_1y)         # expect 50

# Spot-check vs raw Excel:
# Table sheet row 19: AL <- FL = 16,423
flows_2024 |> filter(state_current == "AL", state_1y == "FL")

# Supplemental row for Alabama: 4,522,731 + 437,829 = 4,960,560 stayed-in-AL
flows_2024 |> filter(state_current == "AL", state_1y == "AL")

flows_2024 |> head(5)
```


```{r}
#| label: task3-step2-download

file_url_2023  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                         "tables/geographic-mobility/2023/state-to-state-migration/",
                         "State_to_State_Migration_Table_2023_T13.xlsx")
file_path_2023 <- file.path(data_dir, "State_to_State_Migration_Table_2023.xlsx")

if(!file.exists(file_path_2023)){
    message("Downloading 2023 state-to-state migration file")
    download.file(file_url_2023,
                  destfile = file_path_2023,
                  quiet    = TRUE,
                  mode     = "wb")
}
```

```{r}
#| label: task3-step3-read

# The row/column state ordering in the 2023 sheet matches alphabetical
# with DC inserted after Delaware.
states_with_dc <- c(state.name[1:8], "District of Columbia", state.name[9:50])

# (a) First 6 columns: col D = same residence (didn't move), col F = moved
# within same state. Their sum gives total who stayed inside the state.
# Rows 12-62 cover the 50 states + DC.
stayed_2023 <- read_excel(
        file_path_2023,
        range     = "A12:F62",
        col_names = c("state_current", "pop_est", "pop_moe",
                      "same_res",      "same_res_moe", "same_state")
    ) |>
    mutate(state_current = str_trim(state_current)) |>
    filter(state_current %in% state.name) |>                  # drop DC -> 50 rows
    mutate(population   = as.integer(same_res) + as.integer(same_state),
           state_1y     = state_current,
           year_current = 2023L,
           year_1y      = 2022L) |>
    select(state_current, state_1y, population, year_current, year_1y)

# (b) Bulk flow matrix J:DG. Columns alternate Estimate/MOE, so interleaving
# the state names with NA gives proper names to the Estimate columns and
# placeholder names to the MOE columns, which any_of() then drops.
between_2023 <- read_excel(
        file_path_2023,
        range     = "J12:DG62",
        col_names = vec_interleave(states_with_dc, NA)
    ) |>
    select(any_of(states_with_dc)) |>           # keep 51 Estimate cols
    mutate(state_current = states_with_dc) |>   # tag rows with destination
    pivot_longer(cols      = -state_current,
                 names_to  = "state_1y",
                 values_to = "population") |>
    filter(state_current != "District of Columbia",
           state_1y      != "District of Columbia") |>   # drop DC rows/cols
    filter(state_current != state_1y) |>                 # drop diagonals ("N/A")
    mutate(population   = as.integer(population),
           year_current = 2023L,
           year_1y      = 2022L) |>
    select(state_current, state_1y, population, year_current, year_1y)
```

```{r}
#| label: task3-step4-combine

flows_2023 <- bind_rows(stayed_2023, between_2023) |>
    arrange(state_current, state_1y)

dim(flows_2023)                         # 2500 x 5
n_distinct(flows_2023$state_current)    # 50
n_distinct(flows_2023$state_1y)         # 50

# Spot checks against the raw spreadsheet:
# Row 13 (Alaska) Col J (Alabama) = 204  ->  people in AK who lived in AL a year ago
flows_2023 |> filter(state_current == "Alaska", state_1y == "Alabama")

# Alabama diagonal: 4,482,992 (same res) + 437,745 (moved within AL) = 4,920,737
flows_2023 |> filter(state_current == "Alabama", state_1y == "Alabama")

flows_2023 |> head()
```

```{r}
#| label: task3-function
#| message: false
#| warning: false

library(vctrs)   # for vec_interleave()

state_to_state_migration_2023 <- function(){

    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }

    file_url  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                        "tables/geographic-mobility/2023/state-to-state-migration/",
                        "State_to_State_Migration_Table_2023_T13.xlsx")
    file_path <- file.path(data_dir, "State_to_State_Migration_Table_2023.xlsx")
    if(!file.exists(file_path)){
        message("Downloading 2023 state-to-state migration file")
        download.file(file_url, destfile = file_path, quiet = TRUE, mode = "wb")
    }

    states_with_dc <- c(state.name[1:8], "District of Columbia", state.name[9:50])

    # stayed-in-state diagonal (col D same residence + col F moved within state)
    stayed <- read_excel(
            file_path,
            range     = "A12:F62",
            col_names = c("state_current", "pop_est", "pop_moe",
                          "same_res",      "same_res_moe", "same_state")
        ) |>
        mutate(state_current = str_trim(state_current)) |>
        filter(state_current %in% state.name) |>
        mutate(population   = as.integer(same_res) + as.integer(same_state),
               state_1y     = state_current,
               year_current = 2023L,
               year_1y      = 2022L) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)]) |>
        select(state_current, state_1y, population, year_current, year_1y)

    # between-state flow matrix J:DG; interleave state names with NA so the
    # Estimate columns are named and MOE columns get dropped by any_of()
    between <- read_excel(
            file_path,
            range     = "J12:DG62",
            col_names = vec_interleave(states_with_dc, NA)
        ) |>
        select(any_of(states_with_dc)) |>
        mutate(state_current = states_with_dc) |>
        pivot_longer(cols      = -state_current,
                     names_to  = "state_1y",
                     values_to = "population") |>
        filter(state_current != "District of Columbia",
               state_1y      != "District of Columbia") |>
        filter(state_current != state_1y) |>
        mutate(population   = as.integer(population),
               year_current = 2023L,
               year_1y      = 2022L) |>
        mutate(state_current = state.abb[match(state_current, state.name)],
               state_1y      = state.abb[match(state_1y,      state.name)]) |>
        select(state_current, state_1y, population, year_current, year_1y)

    bind_rows(stayed, between) |>
        arrange(state_current, state_1y)
}
```

```{r}
#| label: task4-function

state_to_state_migration <- function(){
    bind_rows(
        state_to_state_migration_2024(),
        state_to_state_migration_2023()
    ) |>
        arrange(year_current, state_current, state_1y)
}
```

```{r}
#| label: task4-check

flows <- state_to_state_migration()

dim(flows)                                    # expect 5000 x 5
n_distinct(flows$state_current)               # 50
n_distinct(flows$state_1y)                    # 50

# match the four example rows exactly
flows |> filter(state_current == "AL", state_1y == "AL", year_current == 2024)  # 4,960,560
flows |> filter(state_current == "AL", state_1y == "AK", year_current == 2024)  # 387
flows |> filter(state_current == "AL", state_1y == "AL", year_current == 2023)  # 4,920,737
flows |> filter(state_current == "AL", state_1y == "AK", year_current == 2023)  # 780
```



or inline, like this: $x$ is `r x`.

## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 

## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 



------------------------------------------------------------------------

This work ©2025 by  was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
