**This is only a basic Mini-Project Template.**
**Delete this note and customize formatting according to your preferences and**
**the Mini-Project instructions. Your document must start the `---` below**
---
title: "Mini-Project 03 - Who Goes There? US Internal Migration and Implications for Congressional Reapportionment"
author: ""
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Data Import/API Usage/File Parsing in an analysis of Census Data/Migration Flows data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

Remember that code blocks look like this: 

```{r}
#| label: task1-setup
#| message: false
#| warning: false

library(tidyverse)
library(tidycensus)
library(glue)
library(sf)
library(readxl)

# cache tigris shapefile downloads across sessions
options(tigris_use_cache = TRUE)

# project-level data directory (same pattern as mp01, mp02)
data_dir <- file.path("data", "mp03")
if(!dir.exists(data_dir)){
    dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
}
```


```{r}
#| label: task1-pull-acs1
#| message: false

state_pop_file <- file.path(data_dir, "state_population_2015_2024.rds")

if(!file.exists(state_pop_file)){
    
    # ACS-1 was not released for 2020 (pandemic disruption)
    YEARS <- setdiff(2015:2024, 2020)
    
    state_pop_raw <- map(YEARS, function(yy){
        message(glue("Pulling ACS-1 state population for {yy}"))
        get_acs(
            geography = "state", 
            variables = c(total_population = "B01003_001"),
            year      = yy, 
            survey    = "acs1", 
            geometry  = TRUE,    # grab shapefile too
            cache_table = TRUE   # cache the raw table locally
        ) |> 
            mutate(year = yy)
    }) |> bind_rows()
    
    saveRDS(state_pop_raw, state_pop_file)
} else {
    state_pop_raw <- readRDS(state_pop_file)
}
```


```{r}
#| label: task1-tidy

state_pop <- state_pop_raw |> 
    select(GEOID, 
           state_name       = NAME, 
           total_population = estimate, 
           year, 
           geometry)

# quick sanity check: one row per state per year
state_pop |> 
    st_drop_geometry() |> 
    count(year)
```

```{r}
#| label: task2-function
#| message: false
#| warning: false

library(tidyverse)

library(glue)

state_to_state_migration_2024 <- function(){
    
    # 1. Create data directory if it doesn't exist ---------------------------
    data_dir <- file.path("data", "mp03")
    if(!dir.exists(data_dir)){
        dir.create(data_dir, showWarnings = FALSE, recursive = TRUE)
    }
    
    # 2. Download file if not already cached --------------------------------
    file_url  <- paste0("https://www2.census.gov/programs-surveys/demo/",
                        "tables/geographic-mobility/2024/state-to-state-migration/",
                        "State_to_State_Migration_Table_2024_T13.xlsx")
    file_path <- file.path(data_dir, "State_to_State_Migration_Table_2024.xlsx")
    
    if(!file.exists(file_path)){
        message(glue("Downloading 2024 state-to-state migration file from {file_url}"))
        download.file(file_url, destfile = file_path, quiet = TRUE, mode = "wb")
    }
    
    # 3a. "Table" sheet: state-to-state flows -------------------------------
    # Rows 1-8 are merged/metadata headers; actual data starts row 9.
    # 4 cols: current residence, residence 1 yr ago, estimate, margin of error.
    flows_table <- read_excel(
        file_path,
        sheet     = "Table",
        skip      = 8,
        col_names = c("state_current", "state_1y", "estimate", "moe")
    ) |>
        # DC comes in with a trailing space in some rows; trim to be safe
        mutate(state_current = str_trim(state_current),
               state_1y      = str_trim(state_1y)) |>
        # Keep only pairs where BOTH sides are one of the 50 US states.
        # This drops DC, Puerto Rico, Foreign country, U.S. Island Areas,
        # and the footnote rows at the bottom of the sheet.
        filter(state_current %in% state.name,
               state_1y      %in% state.name) |>
        # Drop "non-moving" diagonal rows (value is "X")
        filter(estimate != "X") |>
        # Replace suppressed "N" with "0", then coerce to integer.
        # (dplyr has no replace_values(); if_else does the same job here.)
        mutate(estimate = if_else(estimate == "N", "0", estimate),
               estimate = as.integer(estimate)) |>
        select(state_current,
               state_1y,
               population = estimate) |>
        mutate(year_current = 2024L,
               year_1y      = 2023L)
    
    # 3b. "Supplemental - Current Res" sheet: stayed-in-state totals --------
    # Rows 1-8 headers, row 9 is "United States" (drop), row 10 blank,
    # rows 11-62 are states/DC/PR. Columns we need (by position):
    #   1 = state_current
    #   4 = estimate, "same residence 1 year ago"  (didn't move)
    #   6 = estimate, "movers within same state"   (moved within state)
    # Sum of 4 + 6 = total population that was in the same state 1 yr ago.
    flows_stayed <- read_excel(
        file_path,
        sheet     = "Supplemental - Current Res",
        skip      = 8,
        col_names = FALSE
    ) |>
        select(state_current  = 1,
               same_residence = 4,
               mover_in_state = 6) |>
        mutate(state_current = str_trim(state_current)) |>
        filter(state_current %in% state.name) |>
        mutate(population   = as.integer(same_residence) +
                              as.integer(mover_in_state),
               state_1y     = state_current,
               year_current = 2024L,
               year_1y      = 2023L) |>
        select(state_current, state_1y, population, year_current, year_1y)
    
    # 4. Combine: 2,450 off-diagonal flows + 50 diagonal stay totals = 2,500
    bind_rows(flows_table, flows_stayed) |>
        arrange(state_current, state_1y)
}
```

```{r}
#| label: task2-check

flows_2024 <- state_to_state_migration_2024()

# Shape checks
dim(flows_2024)                  # expect 2500 x 5
n_distinct(flows_2024$state_current)   # expect 50
n_distinct(flows_2024$state_1y)        # expect 50

# Spot-check vs raw Excel:
# Table sheet row 19: Alabama <- Florida = 16,423
flows_2024 |> filter(state_current == "Alabama", state_1y == "Florida")

# Supplemental row for Alabama: 4,522,731 + 437,829 = 4,960,560 stayed-in-AL
flows_2024 |> filter(state_current == "Alabama", state_1y == "Alabama")

flows_2024 |> head(5)
```

or inline, like this: $x$ is `r x`.

## Exploratory Analysis

Code related to Exploratory Data Analysis goes here. This may include
exploratory graphics, instructor-provided exploratory questions, or other
similar elements. 

## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 



------------------------------------------------------------------------

This work ©2025 by  was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #03](https://michael-weylandt.com/STA9750/mini/mini03.html)
