**This is only a basic Mini-Project Template.**
**Delete this note and customize formatting according to your preferences and**
**the Mini-Project instructions. Your document must start the `---` below**
---
title: "Mini-Project 04 - Going for the Gold"
author: "Joshua Eisner"
editor:
    mode: source
format:
    html:
        code-fold: true
---

## Introduction

A suitable introduction goes here. In this mini-project, I intend to demonstrate
my skills at Web Scraping/Data Cleaning/Forecasting in an analysis of Sports Analytics data.

## Data Acquisition and Preparation

Code related to Data Acquisition and Preparation goes here.

Remember that code blocks look like this: 

```{r}
# task1 setup

library(tidyverse)
library(rvest)
library(fs)
library(janitor)
library(necountries)
library(sf)
library(ggplot2)
library(gt)

slow_download <- purrr::slowly(download.file, purrr::rate_delay(pause = 5))

read_olympedia <- function(url, ...) {
  if (stringr::str_starts(url, "/")) url <- stringr::str_remove(url, "/")
  cache_dir <- fs::path("data", "mp04", "olympedia_cache")
  if (!fs::dir_exists(cache_dir)) fs::dir_create(cache_dir, recurse = TRUE)
  cache_path <- fs::path(cache_dir,
                         url |> as.character() |> stringr::str_replace_all("/", "_"))
  if (!fs::file_exists(cache_path)) {
    src_url <- paste0("https://www.olympedia.org/", url)
    slow_download(src_url, cache_path, mode = "wb")
  }
  rvest::read_html(cache_path)
}
```

```{r}
# Task 1 scrape editions
parse_editions_table <- function(table_node, season) {
  if (is.na(table_node)) return(tibble())
  rows <- table_node |> html_elements("tr")

  tibble(
    season      = season,
    olympiad_rn = rows |> html_element("td:nth-of-type(1)") |> html_text2() |> str_trim(),
    edition_id  = rows |> html_element("td:nth-of-type(1) a") |> html_attr("href") |>
                    str_extract("\\d+$") |> as.integer(),
    host_noc    = rows |> html_element("td img") |> html_attr("src") |>
                    str_extract("[A-Z]{3}(?=\\.(png|svg))"),
    year        = rows |> html_element("td:nth-of-type(2)") |> html_text2() |>
                    str_extract("\\b(18|19|20)\\d{2}\\b") |> as.integer()
  ) |>
    mutate(olympiad = suppressWarnings(as.integer(as.roman(olympiad_rn)))) |>
    filter(!is.na(edition_id), !is.na(host_noc), !is.na(year))
}

editions_page <- read_olympedia("editions")

editions <- bind_rows(
  editions_page |>
    html_element(xpath = "(//*[self::h2 or self::h3][normalize-space()='Summer'])[1]/following::table[1]") |>
    parse_editions_table("Summer"),
  editions_page |>
    html_element(xpath = "(//*[self::h2 or self::h3][normalize-space()='Winter'])[1]/following::table[1]") |>
    parse_editions_table("Winter")
) |>
arrange(year, season) |>
  filter(year <= 2024)

nrow(editions)
editions
```

```{r}
# Task 1 scrape-disciplines
get_disciplines <- function(edition_id) {
  page <- read_olympedia(paste0("editions/", edition_id))
  links <- page |> html_elements(paste0("a[href*='/editions/", edition_id, "/sports/']"))
  tibble(
    edition_id      = edition_id,
    discipline_code = links |> html_attr("href") |> str_extract("[^/]+$"),
    discipline_name = links |> html_text2() |> str_trim()
  ) |>
    filter(!is.na(discipline_code), discipline_code != "", str_length(discipline_code) <= 5,
           discipline_name != "") |>
    distinct()
}

disciplines <- editions |>
  mutate(disc = map(edition_id, get_disciplines)) |>
  mutate(disc = map(disc, ~ select(.x, -any_of("edition_id")))) |>
  unnest(disc)


ed_page <- read_olympedia("editions")

ed_page |> html_elements("h1, h2, h3") |> html_text2()

ed_page |> html_elements("table") |> length()
```

```{r}
# Task 1 scrape-medal-tables
get_medal_table <- function(edition_id, discipline_code) {
  page <- read_olympedia(paste0("editions/", edition_id, "/sports/", discipline_code))
  tbl_node <- page |>
    html_element(xpath = "//*[self::h2 or self::h3][normalize-space()='Medal table']/following::table[1]")
  if (is.na(tbl_node)) return(tibble())

  rows <- tbl_node |> html_elements("tr")
  noc <- rows |> html_element("td:nth-of-type(2) a") |> html_attr("href") |>
    str_extract("[A-Z]{3}$")

  raw <- html_table(tbl_node, header = FALSE)
  tibble(
    noc    = noc,
    gold   = suppressWarnings(as.integer(raw[[3]])),
    silver = suppressWarnings(as.integer(raw[[4]])),
    bronze = suppressWarnings(as.integer(raw[[5]])),
    total  = suppressWarnings(as.integer(raw[[6]]))
  ) |>
    filter(!is.na(noc), !is.na(gold))
}


get_medal_table_safe <- possibly(get_medal_table, otherwise = tibble())

medal_tables <- disciplines |>
  mutate(medals = map2(edition_id, discipline_code, get_medal_table_safe)) |>
  unnest(medals)

nrow(medal_tables)
head(medal_tables)
```

```{r}
# Task 1 tidy-final
olympics <- medal_tables |>
  filter(!is.na(noc)) |>
  select(year, season, olympiad, edition_id, host_noc,
         discipline_code, discipline_name,
         noc, gold, silver, bronze, total) |>
  arrange(year, season, edition_id, discipline_code, desc(gold))

glimpse(olympics)
nrow(olympics)
```

```{r}
# task 2 countries-setup



countries_sf <- countries()

# Confirm the 3-letter code column name (likely "a3", but verify)
colnames(countries_sf)
head(countries_sf, 3)

# Find Olympedia codes not in necountries, ranked by medal count

unmatched <- olympics |>
  distinct(noc) |>
  anti_join(countries_sf, by = c("noc" = "iso3"))

unmatched |>
  left_join(olympics |> group_by(noc) |> summarise(total_gold = sum(gold)),
            by = "noc") |>
  arrange(desc(total_gold))
```

```{r}
# Task 2 countries-recode
olympics_clean <- olympics |>
  mutate(noc_iso = case_match(noc,
    "URS" ~ "RUS",  # Soviet Union -> Russia (assignment-required)
    "EUN" ~ "RUS",  # Unified Team 1992 -> Russia (same rationale)
    "FRG" ~ "DEU",  # West Germany -> Germany
    "GDR" ~ "DEU",  # East Germany -> Germany
    "GER" ~ "DEU",  # Olympedia GER -> ISO DEU
    "GRE" ~ "GRC",
    "TCH" ~ "CZE",  # Czechoslovakia -> Czech Republic
    "BUL" ~ "BGR",
    "CRO" ~ "HRV",
    "DEN" ~ "DNK",
    "NED" ~ "NLD",
    "POR" ~ "PRT",
    "RSA" ~ "ZAF",
    "SUI" ~ "CHE",
    "INA" ~ "IDN",
    "IRI" ~ "IRN",
    "PHI" ~ "PHL",
    "ALG" ~ "DZA",
    "CHI" ~ "CHL",
    "MGL" ~ "MNG",
    "SLO" ~ "SVN",
    "LAT" ~ "LVA",
    "URU" ~ "URY",
    "PUR" ~ "PRI",
    "TPE" ~ "TWN",
    "ZIM" ~ "ZWE",
    "ROC" ~ "RUS",
    "BAH" ~ "BHS",
    "NGR" ~ "NGA",
    .default = noc
  ))

# Verify what's left unmatched
olympics_clean |>
  distinct(noc_iso) |>
  anti_join(countries_sf, by = c("noc_iso" = "iso3")) |>
  left_join(olympics_clean |> group_by(noc_iso) |> summarise(total_gold = sum(gold)),
            by = "noc_iso") |>
  arrange(desc(total_gold))
```

```{r}
# Task 2 countries-viz

gold_by_country <- olympics_clean |>
  group_by(noc_iso) |>
  summarise(total_gold = sum(gold, na.rm = TRUE), .groups = "drop")

map_data <- countries_sf |>
  st_set_geometry("polygon") |>
  left_join(gold_by_country, by = c("iso3" = "noc_iso"))

ggplot(map_data) +
  geom_sf(aes(fill = total_gold), color = "white", linewidth = 0.1) +
  scale_fill_viridis_c(
    na.value = "grey90",
    name = "All-time golds",
    labels = scales::label_comma()
  ) +
  labs(
    title = "All-time Olympic gold medal totals by country",
    subtitle = "Modern Olympics, 1896–2026 (Summer + Winter)",
    caption = "Source: Olympedia. Defunct nations recoded to successor states where possible."
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    panel.grid = element_blank(),
    axis.text = element_blank()
  )
```

## Exploratory Analysis

## Task 3

```{r}
# Task 3 preamble

# Filter to summer + season-comparable data for most analyses
olympics_summer <- olympics_clean |> filter(season == "Summer")
```

```{r}
# Task 3 eda-1 USA total medals

usa_total_gold <- olympics_clean |>
  filter(noc == "USA") |>
  summarise(total = sum(gold, na.rm = TRUE)) |>
  pull(total)
usa_total_gold
```
The United States has won `r usa_total_gold` gold medals across all modern Summer and Winter Olympics through 2026 — more than any other country in history.



## Final Insights and Deliverable

Code related to the final deliverable of the assignment goes here. 



------------------------------------------------------------------------

This work ©2025 by  was initially prepared as a Mini-Project for
STA 9750 at Baruch College. More details about this course can be found at
[the course site](https://michael-weylandt.com/STA9750) and instructions for
this assignment can be found at 
[MP #04](https://michael-weylandt.com/STA9750/mini/mini04.html)
